---
title: "PBMC preprocessing"
output: html_notebook
---

In this notebook, we perform preprocessing and QC on our PBMC CITE-seq data. Multimodal clustering and cell type annotation are performed in a follow-up notebook, 2_pbmc_cell_type_annotation.Rmd. The equivalent pipeline for the skin RNA-seq data is executed in 3_skin_preprocessing.Rmd.  

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/paired_skin_and_pbmc/manuscript/code_for_manuscript/"
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))
```

Now, we prepare a Seurat object with one counts matrix for all samples, as well as all necessary metadata.

Prior to preprocessing, we have 44,706 cells total. The sample for patient 515 at 12 months is missing due to quality issues. 

```{r}
################################################################################
## READ & PROCESS META DATA ####################################################
################################################################################

# Read in patient meta data
patient_meta_data <- read.csv(paste0(project_path, "data/patient_meta_data.csv"))
# Remove blank rows
patient_meta_data <- patient_meta_data %>% filter(if_all(everything(), ~ . != "" & !is.na(.)))
# Rename columns as desired
patient_meta_data <- patient_meta_data %>%
  dplyr::rename(
    subject = "subject_id",
    race = "Race",
    ethnicity = "Ethnicity",
    sex = "Sex",
    age_transplant = "Age.at.time.of.ASCT..years.",
    age_onset = "Age.of.onset..years.",
    years_with_disease = "Disease.Duration.at.time.of.ASCT..years.",
    disease_subtype = "Disease.Subtype",
    auto_antibody_pos = "X....Auto.antibody.positivity",
    mRSS = "X....mRSS."
  )

# Turn subject from an integer into a character
patient_meta_data$subject <- as.character(patient_meta_data$subject)

# Separate treatments into individual rows
patient_meta_data_long <- patient_meta_data %>%
  separate_rows(Treatment.prior.to.ASCT, sep = ", ") %>%  # Split treatments into separate rows
  mutate(Treatment.prior.to.ASCT = trimws(Treatment.prior.to.ASCT))  # Trim spaces

# Create columns for each treatment
patient_meta_data_treatments <- patient_meta_data_long %>%
  mutate(value = TRUE) %>%  # Add a value column to indicate presence
  pivot_wider(names_from = Treatment.prior.to.ASCT,
              values_from = value,
              values_fill = list(value = FALSE))  # Fill with FALSE if not present

# Merge back with original metadata
patient_meta_data <- patient_meta_data %>%
  select(subject, race, ethnicity,sex, age_transplant, age_onset, years_with_disease, disease_subtype) %>%
  left_join(patient_meta_data_treatments, by = "subject")

# Remove major manifestations column
patient_meta_data$Major.manifestations <- NULL

################################################################################
## READ IN COUNTS MATRICES #####################################################
################################################################################
#
# # Get paths to cellranger multi output
# cellranger_paths <- list.files(paste0(project_path, "data/cellranger_output/PBMC/cellranger_multi"), full.names = TRUE)
# 
# # Get sample names
# samp_names <- list.files(paste0(project_path, "data/cellranger_output/PBMC/cellranger_multi"))
# 
# # Read 10x output
# cellranger_output <- lapply(cellranger_paths, Read10X)
# 
# # GEX all in list
# gex_list <- lapply(cellranger_output, function(x){x$`Gene Expression`})
# 
# # PLX all in list
# plx_list <- lapply(cellranger_output, function(x){x$`Antibody Capture`})
# 
# # Save as RDS files
# mapply(saveRDS, gex_list, paste0(project_path, "data/RDS_files/counts/PBMC/gex/", samp_names, "_gex.RDS"))
# mapply(saveRDS, plx_list, paste0(project_path, "data/RDS_files/counts/PBMC/plx/", samp_names, "_plx.RDS"))

# Read in RDS files
gex_paths <- list.files(paste0(project_path, "data/RDS_files/counts/PBMC/gex"), pattern = "\\.RDS$", full.names = TRUE)
plx_paths <- list.files(paste0(project_path, "data/RDS_files/counts/PBMC/plx"), pattern = "\\.RDS$", full.names = TRUE)

# Read in counts
gex_list <- lapply(gex_paths, readRDS)
plx_list <- lapply(plx_paths, readRDS)

# Remove underscores from antibody names (Seurat doesn't allow)
plx_list <- lapply(plx_list, function(mat){rownames(mat) <- gsub("_", "-", rownames(mat)); mat})

# Name elements according to corresponding sample
gex_sample_ids <- sub("_gex\\.RDS$", "", basename(gex_paths))
names(gex_list) <- gex_sample_ids

plx_sample_ids <- sub("_plx\\.RDS$", "", basename(plx_paths))
names(plx_list) <- plx_sample_ids

# If sample IDs match (as they should), store in sample_ids object
if (all(gex_sample_ids == plx_sample_ids)) {
  sample_ids <- gex_sample_ids
} else {stop("Sample IDs from gex and plx do not match.")}

################################################################################
## CREATE SEURAT OBJECT ########################################################
################################################################################

# We need V5 to use JoinLayers later (but will have to create a v3 object
# to use with SeuratDisk)
options(Seurat.object.assay.version = "v5")

#### GEX #######################################################################
# List of Seurat objects (1 for each sample)
scler_gex_list <- lapply(gex_list,
                         CreateSeuratObject,
                         project = "scl-CITE-Seq",
                         assay = "RNA")


# Combine into one Seurat object
scler_gex_obj <- merge(scler_gex_list[[1]],
                        scler_gex_list[-1],
                        add.cell.ids = sample_ids,
                        project = "scl-CITE-Seq")
# Join layers
scler_gex_joined <- JoinLayers(scler_gex_obj)

#### PLX #######################################################################
# List of Seurat objects (1 for each sample)
scler_plx_list <- lapply(plx_list,
                         CreateSeuratObject,
                         project = "scl-CITE-Seq",
                         assay = "ADT")


# Combine into one Seurat object
scler_plx_obj <- merge(scler_plx_list[[1]],
                        scler_plx_list[-1],
                        add.cell.ids = sample_ids,
                        project = "scl-CITE-Seq")

# Join layers
scler_plx_joined <- JoinLayers(scler_plx_obj)

#### PUT GEX AND PLX INTO ONE OBEJECT ##########################################
scler_pbmc_all <- scler_gex_joined
scler_pbmc_all[["ADT"]] <- scler_plx_joined[["ADT"]]

################################################################################
## ADD META DATA TO SEURAT OBJECTS #############################################
################################################################################

### (a) sample ID ##############################################################
scler_pbmc_all$sample_id <- sub("_(?!.*_).*",
                                   "",
                                   rownames(scler_pbmc_all@meta.data),
                                   perl = TRUE)

### (b) subject ################################################################
scler_pbmc_all@meta.data <- scler_pbmc_all@meta.data %>%
  mutate(
    subject = case_when(
      grepl("^H", sample_id) ~ sample_id,  # Healthy controls start with 'H'
      TRUE ~ sub("^[^_]*_", "", sample_id)  # Everyone else: remove prefix before underscore
    )
  )

### (c) LIBRARY ################################################################
scler_pbmc_all@meta.data <- scler_pbmc_all@meta.data %>%
  mutate(
    LIBRARY = case_when(
      grepl("H", subject) ~ "H", # Healthy controls are part of 'H' library
      TRUE ~ subject # Each patient has is a library
    )
  )

### (d) health #################################################################
scler_pbmc_all@meta.data <- scler_pbmc_all@meta.data %>%
  mutate(
    health = case_when(
      grepl("H", sample_id) ~ "healthy",
      TRUE ~ "SSc"
    )
  )

### (e) group
scler_pbmc_all@meta.data <- scler_pbmc_all@meta.data %>%
  mutate(
    group = case_when(
      health == "healthy" ~ "healthy",
      grepl("baseline", sample_id) ~ "baseline",
      grepl("six", sample_id) ~ "six",
      grepl("twelve", sample_id) ~ "twelve",
      grepl("twen4", sample_id) ~ "twen4"
    ),
    group = factor(group, levels = c("healthy", "baseline", "six", "twelve", "twen4"))
  )

### (f) Add timepoint as a numeric variable for patients
scler_pbmc_all@meta.data <- scler_pbmc_all@meta.data %>%
  mutate(
    timepoint = case_when(
      group == "healthy" ~ NA,
      group == "baseline" ~ 0,
      group == "six" ~ 6,
      group == "twelve" ~ 12,
      group == "twen4" ~ 24
    )
  )


### (g) Add other patient meta data (while retaining metadata rownames)
scler_pbmc_all@meta.data <- scler_pbmc_all@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  left_join(patient_meta_data, by = "subject") %>%
  tibble::column_to_rownames("cell_id")

################################################################################
## RENAME CELLS TO MATCH LIBRARY  ##############################################
################################################################################

colnames(scler_pbmc_all) <- colnames(scler_pbmc_all) %>%
  gsub("-1$", "", .) %>%                              # remove trailing -1
  gsub(".*_([0-9]+)_", "\\1-", .) %>%                 # patient: extract any numeric ID and add dash
  gsub("H[0-9]+_", "H-", .)                           # healthy: H1_, H2_ → H-

################################################################################
## SAVE SERURAT OBJECT #########################################################
################################################################################

# We need to turn this V5 Seurat object into a V3 Seurat object to be compatible
# with SeuratDisk
options(Seurat.object.assay.version = "v3")
# Create object with RNA and metadata
scler_pbmc <- CreateSeuratObject(counts = scler_pbmc_all@assays$RNA$counts,
                                 meta.data = scler_pbmc_all@meta.data,
                                 project = scler_pbmc_all@project.name)
# Add ADT
scler_pbmc[["ADT"]] <- CreateAssayObject(counts = scler_pbmc_all[["ADT"]]$counts)

# Save as H5
SaveH5Seurat(scler_pbmc,
             paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_1"))
```

Now, we begin preprocessing

1) Multimodal multiplet detection with COMPOSITE.

First, we extract the RNA and ADT counts matrices, filter out genes with > 50% zeroes, and save both matrices as mtx files.

```{r}
# Load in most recent Seurat object
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_1.h5seurat"))

# Extract RNA and ADT
rna <- scler_pbmc@assays$RNA@counts
adt <- scler_pbmc@assays$ADT@counts

# Filter out genes with > 50% zeroes.
# Starting with 38606 genes
# Calculate proportion of zeros for each gene
rna_prop_zero <- (rna == 0) %>% rowMeans()
# Filter out genes with more than 50% zeros
rna <- rna[names(which(rna_prop_zero < 0.5)), ]

# Write to mtx files
rna %>% writeMM(file = paste0(project_path, "data/composite/PBMC/RNA.mtx"))
adt %>% writeMM(file = paste0(project_path, "data/composite/PBMC/ADT.mtx"))
```

Composite is implemented in Python in the file, composite_pbmc.py.

See https://github.com/CHPGenetics/COMPOSITE

We import the multiplet predictions as a CSV file.

We detect 10,568 multiplets (23.6%).

After removing multiplets, 34,138 cells remain.

```{r}
# Import sccomposite multiplet predictions
composite_path <- paste0(project_path, "data/composite/PBMC/multiplet_prediction.csv")

# Add sccomposite labels to Seurat object
scler_pbmc$multiplet_label <- read.csv(composite_path) %>% pull(multiplet_classification)

# How many multiplets? 10,568
scler_pbmc$multiplet_label %>% sum()

# What percentage of all cells are multiplets? 23.6%
(scler_pbmc$multiplet_label %>% sum()) / (scler_pbmc$multiplet_label %>% length()) * 100

# Remove multiplets 
scler_pbmc <- subset(scler_pbmc, subset = multiplet_label == 0)

# How many remaining cells? 34,138
scler_pbmc$multiplet_label %>% length()

# Remove multiplet_label (all zero now)
scler_pbmc$multiplet_label <- NULL

# Save as H5
SaveH5Seurat(scler_pbmc,
             paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_2"))
```


2) Ambient RNA removal with SoupX. Since this step addresses technical noise, we run it separately for each library (subject), so we need the output from cellranger count, before we demultiplexed with cellranger multi. 

```{r}
# Path where cellranger_count outputs are saved
path_to_cellranger_count <- paste0(project_path, "data/cellranger_output/PBMC/cellranger_count/")

# Load (demultiplexed) Seurat object
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_2.h5seurat"))

# Subjects
subjects <- list.files(paste0(path_to_cellranger_count, "raw"))

# Load data for raw and filtered wells
raw_wells <- lapply(subjects, function(subject){Read10X(paste0(path_to_cellranger_count, "raw/", subject))})
filtered_wells <- lapply(subjects, function(subject){Read10X(paste0(path_to_cellranger_count, "filtered/", subject))})

# Name accordingly
names(raw_wells) <- names(filtered_wells) <- subjects

# # Run soup for each well individually
# soup_wells <- mapply(run_soup,
#                     raw_counts = raw_wells,
#                     filtered_counts = filtered_wells,
#                     subject = gsub("^p", "", subjects),
#                     SIMPLIFY = F)
# # Save as RDS
# saveRDS(soup_wells, paste0(project_path, "data/RDS_files/soup_wells/PBMC_soup.RDS"))

# Open RDS of SoupX output
soup_wells <- readRDS(paste0(project_path, "data/RDS_files/soup_wells/PBMC_soup.RDS"))

# Let's look at fraction of contamination in each library
# Ranges from 1.3-2.4%

for (subject in subjects) {
  prop_contamination <- soup_wells[[subject]]$sc$metaData$rho
  print(paste0(subject, ": ", unique(prop_contamination)))
}

# Change dashes to underscores in scler_pbmc cell names
colnames(scler_pbmc) <- gsub("-", "_", colnames(scler_pbmc))

# Merge into one big matrix
counts_adj <- do.call(cbind, lapply(soup_wells, function(x){x$adj_counts}))

# Are there any cells filtered out by cellranger count that were "rescued" by cellranger multi?
extra_cells <- setdiff(colnames(scler_pbmc), colnames(counts_adj))

# There are only three cells that were "rescued". We can just remove them.
scler_pbmc <- subset(scler_pbmc, cells = setdiff(Cells(scler_pbmc), extra_cells))

# Now, subset counts_adj to only include cells in scler_pbmc
counts_adj <- counts_adj[, Cells(scler_pbmc)]

# Adjust original object
scler_pbmc[["RNA"]]$counts <- counts_adj

# Save as H5
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_3"))
```

3) Quality control based on RNA and ADT features

Now, we perform some basic quality control, removing cells which express outlier quantities of unique genes, unique antibodies, total transcripts, total antibodies, percentage of mitochondrial DNA, or percentage of ribosomal DNA than the other sampled cells.

After this filtering step, we retain 29,805 cells.

```{r}
# ################################################################################
# ## LOAD SEURAT OBJECT #########################################################
# ################################################################################

scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_3.h5seurat"))

# ################################################################################
# ## FILTERING AND QC ############################################################
# ################################################################################

# 1. RNA #######################################################################

# Add metadata for percentage of mitochondrial and ribosomal DNA
scler_pbmc[["percent.mt"]] <- PercentageFeatureSet(scler_pbmc, pattern = "^MT-", assay = "RNA")
scler_pbmc[["percent.ribo"]] <- PercentageFeatureSet(scler_pbmc, pattern = "^RP[SL]", assay = "RNA")

# Now, look check violin plots for:
#    - nFeature_RNA: number of unique genes expressed per cell
#    - nCount_RNA: total number of RNA transcripts detected per cell
#    - percent.mt: percentage of reads that map to the mitochondrial genome, and
#    - percent.ribo: percentage of reads that map to the ribosomal genome.
VlnPlot(scler_pbmc, features = c("nFeature_RNA"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("nCount_RNA"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("percent.mt"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("percent.ribo"), pt.size = 0, group.by = "sample_id")

# Identify cells as "outliers" if median absolute deviation for these metrics >= 3
# Adaptive thresholds are calculated separately for each library
qc.nCount_RNA <- isOutlier(scler_pbmc$nCount_RNA, log = TRUE, type = "both", batch = scler_pbmc$LIBRARY)
qc.nFeature_RNA  <- isOutlier(scler_pbmc$nFeature_RNA, log = TRUE, type="both", batch = scler_pbmc$LIBRARY)
qc.percent.mt <- isOutlier(scler_pbmc$percent.mt, log = FALSE, type = "higher", batch = scler_pbmc$LIBRARY)
qc.percent.ribo <- isOutlier(scler_pbmc$percent.ribo, log = FALSE, type = "higher", batch = scler_pbmc$LIBRARY)

# Indicator for whether a cell is an outlier by at least one of this metrics
scler_pbmc$outlier_status <- (qc.nCount_RNA | qc.nFeature_RNA | qc.percent.mt | qc.percent.ribo)

# How many cells do we have before filtering?
print(paste0("cells before filtering: ", ncol(scler_pbmc))) # 34135

# Remove outliers
scler_pbmc <- subset(scler_pbmc,
                     outlier_status == F)
# Remove outlier_status variable
scler_pbmc$outlier_status <- NULL

# How many cells do we have after filtering?
print(paste0("cells after filtering: ", ncol(scler_pbmc))) # 30405

# Check plots again
VlnPlot(scler_pbmc, features = c("nFeature_RNA"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("nCount_RNA"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("percent.mt"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("percent.ribo"), pt.size = 0, group.by = "sample_id")

# 2. ADT #######################################################################

VlnPlot(scler_pbmc, features = c("nFeature_ADT"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("nCount_ADT"), pt.size = 0, group.by = "sample_id")

# Identify cells as "outliers" if median absolute deviation for these metrics >= 3
# Adaptive thresholds are calculated separately for each library
qc.nCount_ADT <- isOutlier(scler_pbmc$nCount_ADT, log = FALSE, type = "both", batch = scler_pbmc$LIBRARY)
qc.nFeature_ADT  <- isOutlier(scler_pbmc$nFeature_ADT, log = FALSE, type="both", batch = scler_pbmc$LIBRARY)

# Indicator for whether a cell is an outlier by at least one of this metrics
scler_pbmc$outlier_status <- (qc.nCount_ADT | qc.nFeature_ADT)

# How many cells do we have before filtering?
print(paste0("cells before filtering: ", ncol(scler_pbmc))) # 30405

# Remove outliers
scler_pbmc <- subset(scler_pbmc,
                     outlier_status == F)
# Remove outlier_status variable
scler_pbmc$outlier_status <- NULL

# How many cells do we have after filtering?
print(paste0("cells after filtering: ", ncol(scler_pbmc))) # 29805

# Check plots again
VlnPlot(scler_pbmc, features = c("nFeature_ADT"), pt.size = 0, group.by = "sample_id")
VlnPlot(scler_pbmc, features = c("nCount_ADT"), pt.size = 0, group.by = "sample_id")
```

Isotope controls do not target any specific antigen, and are used to identify "sticky" cells. We filter out sticky cells that express too many of these non-specific antibodies. We retain 28,943 cells after this step.

```{r}
# Extract ADT counts matrix
ADT_counts <- scler_pbmc@assays$ADT@counts
# Extract isotype control names
isotype_ctls <- rownames(ADT_counts)[str_detect(rownames(ADT_counts), fixed("Iso", ignore_case = TRUE))]
# Extract counts for isotype controls
isotype_ctl_data <- as.matrix(ADT_counts[isotype_ctls,])

# Add metadata for each isotype control
for(i in 1:length(rownames(isotype_ctl_data))){
  isotype_name <- rownames(isotype_ctl_data)[[i]]
  scler_pbmc <- AddMetaData(scler_pbmc,
                   unlist(isotype_ctl_data[isotype_name,]),
                   sprintf("%s",str_replace_all(isotype_name, "-", "_")))
}

# Add metadata for maximum isotype control percentage (highest percentage from the isotype control % features)
scler_pbmc <- AddMetaData(scler_pbmc, apply(isotype_ctl_data, 2, max), "isotype_ctl_max")

# Visualize with violin plot
VlnPlot(scler_pbmc, features = c("isotype_ctl_max"), pt.size = 0, group.by = "sample_id")

# How many cells do we have before filtering?
print(paste0("cells before filtering: ", ncol(scler_pbmc))) # 29805

# Find outliers usig adapative thresholds by library
qc.isotype_ctl_max <- isOutlier(scler_pbmc$isotype_ctl_max, log = FALSE, type = "higher", batch = scler_pbmc$LIBRARY)
# Add variable
scler_pbmc$outlier_status <- qc.isotype_ctl_max
# Remove outliers
scler_pbmc <- subset(scler_pbmc, outlier_status == F)
# Remove variable
scler_pbmc$outlier_status <- NULL

# How many cells do we have after filtering?
print(paste0("cells after filtering: ", ncol(scler_pbmc))) # 28943

# Check isotype_ctl_max violin plot again:
VlnPlot(scler_pbmc, features = c("isotype_ctl_max"), pt.size = 0, group.by = "sample_id")
# Save Seurat object
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_4"))

```

We'll save these thresholds in a text file for future reference.

```{r}
qc_features <- list("nCount_RNA" = qc.nCount_RNA,
                 "nFeature_RNA" = qc.nFeature_RNA,
                 "percent.mt" = qc.percent.mt,
                 "percent.ribo" = qc.percent.ribo,
                 "nCount_ADT" = qc.nCount_ADT,
                 "nFeature_ADT" = qc.nFeature_ADT,
                 "isotype_ctl_max" = qc.isotype_ctl_max)
threshold_list <- lapply(qc_features, attr, "thresholds")

sink(paste0(project_path, "data/qc/pbmc_thresholds.txt"))
threshold_list
sink
```

4) Gene filtering

We will filter genes that are lowly expressed as well. Before filtering, we have 38,606 genes.

We only consider expression in patients, as we do not include healthy controls in our model.

```{r}
# Load Seurat object
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_4.h5seurat"))

# Take subset of just patients
patient_subset <- subset(scler_pbmc, health == "SSc")

# Aggregate expression by sample
pb_patient <- PseudobulkExpression(patient_subset,
                     group.by = "sample_id",
                     return.seurat = F,
                     layer = "counts",
                     method = "aggregate") %>%
    .[["RNA"]] %>% suppressWarnings()
# Convert to CPM
pb_patient_cpm <- sweep(pb_patient, 2, colSums(pb_patient), "/") * 10e6

# Get timepoint vector
agg_timepoint <- sapply(colnames(pb_patient), function(s) {
    # extract timepoint from sample name
    strsplit(s, "-")[[1]][1]
})

# Get a list of genes that has a count >= 3 for at least two patients at at least one time point.
baseline_keep <- (pb_patient_cpm[, agg_timepoint == "baseline"] > 3) %>% rowSums() >= 2
six_keep <- (pb_patient_cpm[, agg_timepoint == "six"] > 3) %>% rowSums() >= 2
twelve_keep <- (pb_patient_cpm[, agg_timepoint == "twelve"] > 3) %>% rowSums() >= 2
twen4_keep <- (pb_patient_cpm[, agg_timepoint == "twen4"] > 3) %>% rowSums() >= 2
any_keep <- baseline_keep | six_keep | twelve_keep | twen4_keep
keep_genes <- names(any_keep)[any_keep]

# How many genes before filtering? 38743
print(paste0("Genes before filtering: ", nrow(scler_pbmc)))

# Subset RNA assay
scler_pbmc[["RNA"]] <- subset(scler_pbmc[["RNA"]], features = keep_genes)

# How many genes after filtering? 18766
print(paste0("Genes after filtering: ", nrow(scler_pbmc)))

# Save H5
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_5"))
```

After filtering, we have 18,766 genes.

5) Normalize & scale RNA data

We've now denoised and filtered the RNA data. Lets normalize and scale it.

```{r}
# Load Seurat object
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_5.h5seurat"))

# Set default assay to RNA (we will address ADT layer later)
DefaultAssay(scler_pbmc) <- "RNA"
# Normalize, find variable features (default 2000), and scale 
scler_pbmc <- NormalizeData(scler_pbmc) %>%
  FindVariableFeatures() %>%
  ScaleData(features = rownames(scler_pbmc))

# Save as H5
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_6"))
```

6) Denoise and normalize ADT data

Now, let's denoise and normalize the ADT data with DSB. We set denoise.counts = FALSE since our signal is weak and our ADT data is not bimodal. We also scale the data. 

For this step, we use the output from cellranger count, before we demultiplexed the data with cellranger multi.

```{r}
# Path where cellranger_count outputs are saved
path_to_cellranger_count <- paste0(project_path, "data/cellranger_output/PBMC/cellranger_count/")

# Load (demultiplexed) Seurat object
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_6.h5seurat"))

# Subjects
subjects <- list.files(paste0(path_to_cellranger_count, "raw"))

# Load data for raw and filtered wells
raw_wells <- lapply(subjects, function(subject){Read10X(paste0(path_to_cellranger_count, "raw/", subject))})
filtered_wells <- lapply(subjects, function(subject){Read10X(paste0(path_to_cellranger_count, "filtered/", subject))})

# Name accordingly
names(raw_wells) <- names(filtered_wells) <- subjects

# Look at distribution of droplets for each well
dsb_plots <- mapply(plot_dsb, raw_well, filtered_well,
       SIMPLIFY = F)
# Name accordingly
names(dsb_plots) <- subjects

# Get data classifying cells as background/cell
dsb_data <- mapply(dsb_md, raw_well, filtered_well,
                   SIMPLIFY = F)
# Name accordingly
names(dsb_data) <- subjects

# Get cutoffs using scater

# RNA size
rna.size_cutoffs <- lapply(subjects,
                           function(subject){
                             isOutlier((dsb_data[[subject]] %>% filter(drop.class == "cell") %>% pull(rna.size)),
          log = TRUE,
          type = "both") %>% attr("thresholds")
          })
# Name accordingly
names(rna.size_cutoffs) <- subjects
# Get lower and upper cutoffs
rna.lower <- lapply(rna.size_cutoffs, function(x){x[1]}); rna.higher <- lapply(rna.size_cutoffs, function(x){x[2]})

# ADT size
prot.size_cutoffs <- lapply(subjects,
                           function(subject){
                             isOutlier((dsb_data[[subject]] %>% filter(drop.class == "cell") %>% pull(prot.size)),
          log = TRUE,
          type = "both") %>% attr("thresholds")
          })
# Name accordingly
names(prot.size_cutoffs) <- subjects
# Get lower and upper cutoffs
prot.lower <- lapply(prot.size_cutoffs, function(x){x[1]}); prot.higher <- lapply(prot.size_cutoffs, function(x){x[2]})


# Pct mt (mitochondrial DNA) max
mt.prop_cutoffs <- lapply(subjects,
                           function(subject){
                             isOutlier((dsb_data[[subject]] %>% filter(drop.class == "cell") %>% pull(mt.prop)),
          log = FALSE,
          type = "higher") %>% attr("thresholds")
          })
# Name accordingly
names(mt.prop_cutoffs) <- subjects
# Get lower and upper cutoffs
mt.prop.higher <- lapply(mt.prop_cutoffs, function(x){x[2]})

# Get libraries
libraries <- unique(scler_pbmc$LIBRARY)
# List of Seurat objects subset by library
seurat_subsets <- lapply(libraries, function(lib){subset(scler_pbmc, LIBRARY == lib)})

# Run DSB for each well/library
adt_adj <- mapply(run_dsb,
                  raw_well, filtered_well, seurat_subsets, libraries,
                  prot.size.min = prot.lower,
                  prot.size.max = prot.higher,
                  rna.size.min = rna.lower,
                  rna.size.max = rna.higher,
                  mt.prop.max = mt.prop.higher,
                  SIMPLIFY = F,
                  MoreArgs = list(denoise.counts = FALSE,
                                  define.pseudocount = TRUE,
                                  pseudocount.use = 1))

# Put adjusted ADT into one big matrix
adt_adj_mat <- do.call(cbind, adt_adj)

# Make sure order of rows and columns matches Seurat object
adt_adj_mat_sorted <- adt_adj_mat[rownames(scler_pbmc[["ADT"]]), colnames(scler_pbmc[["ADT"]])]

# Check dimensions and row/column names
stopifnot(all(rownames(adt_adj_mat_sorted) == rownames(scler_pbmc[["ADT"]])))
stopifnot(all(colnames(adt_adj_mat_sorted) == colnames(scler_pbmc[["ADT"]])))

# Update Seurat object with DSB normalized matrix
scler_pbmc[["DSB"]] <- Seurat::CreateAssayObject(data = adt_adj_mat_sorted)

# Set DSB as default layer
DefaultAssay(scler_pbmc) <- "DSB"

# Scale DSB data
# Get isotype controls
isotype_ctls <- rownames(scler_pbmc[["DSB"]])[str_detect(rownames(scler_pbmc[["DSB"]]), fixed("Iso", ignore_case = TRUE))]
# Get features from DSB layer (excluding isotype controls)
dsb_features <- setdiff(rownames(scler_pbmc[["DSB"]]), isotype_ctls)
# Set features as variable features
VariableFeatures(scler_pbmc) <- dsb_features
# Scale
scler_pbmc <- scler_pbmc %>%
  ScaleData()

# Save Seurat object as H5
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_7"))
```

Ok, so both the RNA and ADT are denoised, normalized, and scaled.

7) Batch integration

Now, we integrate the RNA data using harmony and the ADT data using RPCA to account for batch effects.

```{r}
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_7.h5seurat"))

# Data integration

# (1) RNA - Run Harmony
# Set default assay
DefaultAssay(scler_pbmc) <- "RNA"
# Run PCA
scler_pbmc <- RunPCA(scler_pbmc, reduction.name = "pca_rna")
# Set dimensions
rna_dims <- 1:30
# Integrate with harmony
scler_pbmc <- RunHarmony(scler_pbmc, group.by.vars = "LIBRARY", reduction.use = "pca_rna", dims.use = rna_dims, verbose = TRUE)

# (2) ADT - Run RPCA
# Set default assay
DefaultAssay(scler_pbmc) <- "DSB"
# Integrate with RPCA (Seurat integration)
scler_pbmc <- run_rpca(scler_pbmc)

# Save integrated Seurat object
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/PBMC/scler_pbmc_8"))
```

