---
title: "2: PBMC cell type annotation"
output: html_notebook
---

This notebook is a follow-up to 1_pbmc_preprocessing.Rmd. After preprocessing our PBMC data, we are ready to perform multimodal clustering and cell type annotation.

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater", "RColorBrewer")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))

# Function to update cell_types
update_cell_types <- function(df, cluster, cell_type = NA) {
  df[df$seurat_clusters == cluster, "cell_type"] <- cell_type
  return(df)
}
```

Now, we load in the Seurat object we generated in the previous notebook.

```{r}
# # Load Seurat object
# scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_8.h5seurat"))
# 
# # RNA harmony reduction is ready to be used with WNN.
# 
# # Now, set up integrated DSB values to use in WNN analysis.
# DefaultAssay(scler_pbmc) <- "DSB_integrated"
# 
# # Scale and run PCA on integrated DSB data
# scler_pbmc <- scler_pbmc %>%
#   ScaleData() %>%
#   RunPCA(reduction.name = 'pca_dsb_int', verbose = FALSE)
# 
# # Run WSNN
# scler_pbmc <- FindMultiModalNeighbors(
#   scler_pbmc, reduction.list = list("harmony", "pca_dsb_int"),
#   dims.list = list(1:30, 1:18),
#   verbose = FALSE,
#   k.nn = 20
# )
# 
# # Run UMAP
# scler_pbmc <- RunUMAP(scler_pbmc, 
#                       nn.name = "weighted.nn", 
#                       reduction.name = "wnn.umap", 
#                       reduction.key = "wnnUMAP_")
# 
# 
# # Cluster using SLM algorithm (3)
# scler_pbmc <- FindClusters(scler_pbmc, graph.name = "wsnn",
#                   algorithm = 3,
#                   resolution = 0.3,
#                   random.seed = 1990)
# 
# # Save as H5
# SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/scler_pbmc_9"))
```

Now, we use Seurat's FindMarkers function to get differentially expressed genes and proteins for each cluster.

```{r}
# # # Load data
# scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_9.h5seurat"))
# # 
# # # Take a look
# DimPlot(scler_pbmc, label = T, repel = T)
# # 
# # # Get markers by cluster
# # 
# # # # RNA markers
# # # rna_de <- FindAllMarkers(scler_pbmc, assay = "RNA", only.pos = T)
# # # # Save as RDS
# # # saveRDS(rna_de, paste0(project_path, "data/RDS_files/markers/rna_de.RDS"))
# # # Read RDS
# rna_de <- readRDS(paste0(project_path, "data/RDS_files/markers/rna_de.RDS"))
# # 
# # # # ADT markers
# # # dsb_de <- FindAllMarkers(scler_pbmc, assay = "DSB", only.pos = T)
# # # # Save as RDS
# # # saveRDS(dsb_de, paste0(project_path, "data/RDS_files/markers/dsb_de.RDS"))
# # # Read RDS
# dsb_de <- readRDS(paste0(project_path, "data/RDS_files/markers/dsb_de.RDS"))
```

Now we annotate each cluster based on canonical RNA and protein markers. 

```{r}
# Data frame to store cell types
cell_types <- data.frame(seurat_clusters = levels(scler_pbmc$seurat_clusters),
                         cell_type = NA)
################################################################################
#  CLUSTER 0                                                                   #
#  TYPE: MONOCYTE                                                        #
#  RNA MARKERS: CD14, LYZ                                                      #
#  ADT MARKERS:                                                                #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 0); display_top(dsb_de, 0)
# # Look at CD14
# rna_de %>% filter(gene == "CD14" & cluster == 0)
# Update cell_types
cell_types <- update_cell_types(cell_types, 0, cell_type = "monocyte")
################################################################################
#  CLUSTER 1                                                                   #
#  TYPE: B CELL                                                                #
#  RNA MARKERS: CD79A, MS4A1, CD79B, IGHM                                      #
#  ADT MARKERS: IgM                                                            #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 1); display_top(dsb_de, 1)
# Update cell_types
cell_types <- update_cell_types(cell_types, 1, cell_type = "B cell")
################################################################################
#  CLUSTER 2                                                                   #
#  TYPE: CD4+ T CELL                                                           #
#  RNA MARKERS: CD3E, CD3D, CD3G, CD4                                          #
#  ADT MARKERS: CD2, CD7                                                       #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 2); display_top(dsb_de, 2)
# # Look at CD4
# rna_de %>% filter(cluster == 2 & gene == "CD4")
# Update cell_types
cell_types <- update_cell_types(cell_types, 2, cell_type = "CD4+ T cell")
################################################################################
#  CLUSTER 3                                                                   #
#  TYPE: MONOCYTE                                                        #
#  RNA MARKERS: FCGR3A                                                         #
#  ADT MARKERS: CD16                                                           #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 3); display_top(dsb_de, 3)
# # Look at FCGR3A (CD16)
# rna_de %>% filter(cluster == 3 & gene == "FCGR3A")
# Update cell_types
cell_types <- update_cell_types(cell_types, 3, cell_type = "monocyte")
################################################################################
#  CLUSTER 4                                                                   #
#  TYPE: NATURAL KILLER                                                        #
#  RNA MARKERS: GNLY, NKG7, GZMA, GZMB, KLRB1, KLRG1                           #
#  ADT MARKERS: CD56, CD16                                                     #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 4); display_top(dsb_de, 4)
# Update cell_types
cell_types <- update_cell_types(cell_types, 4, cell_type = "natural killer")
################################################################################
#  CLUSTER 5                                                                   #
#  TYPE: CD8+ T CELL                                                           #
#  RNA MARKERS: CD3E, CD3D, CD8B, CD8A, CD7                                    #
#  ADT MARKERS: CD2, CD8, CD3, CD5, CD7                                        #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 5); display_top(dsb_de, 5)
# Update cell_types
cell_types <- update_cell_types(cell_types, 5, cell_type = "CD8+ T cell")
################################################################################
#  CLUSTER 6                                                                   #
#  TYPE: NATURAL KILLER                                                        #
#  RNA MARKERS: GNLY, KLRB1, NKG7, GZMA, GZMB, NCAM1                           #
#  ADT MARKERS: CD56                                                           #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 6); display_top(dsb_de, 6)
# # Look at NCAM1
# rna_de %>% filter(cluster == 6 & gene == "NCAM1")
# # Look at CD56
# dsb_de %>% filter(cluster == 6 & gene == "anti-human-CD56")
# Update cell_types
cell_types <- update_cell_types(cell_types, 6, cell_type = "natural killer")
################################################################################
#  CLUSTER 7                                                                   #
#  TYPE: MONOCYTE                                                        #
#  RNA MARKERS: LYZ, CD14                                                      #
#  ADT MARKERS:                                                                #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 7); display_top(dsb_de, 7)
# Update cell_types
cell_types <- update_cell_types(cell_types, 7, cell_type = "monocyte")
################################################################################
#  CLUSTER 8                                                                   #
#  TYPE: B CELL                                                                #
#  RNA MARKERS: CD79A, IGHM, MS4A1, IGHD                                       #
#  ADT MARKERS: CD19, IgM, CD20, CD21, IgD                                     #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 8); display_top(dsb_de, 8)
# Update cell_types
cell_types <- update_cell_types(cell_types, 8, cell_type = "B cell")
################################################################################
#  CLUSTER 9                                                                   #
#  TYPE: mDC                                                                   #
#  RNA MARKERS: CD1C                                                           #
#  ADT MARKERS: CD1C                                                           #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 9); display_top(dsb_de, 9)
# # Look at CD1C
# rna_de %>% filter(cluster == 9 & gene == "CD1C"); dsb_de %>% filter(cluster == 9 & gene == "anti-human-CD1c")
# Update cell_types
cell_types <- update_cell_types(cell_types, 9, cell_type = "mDC")
################################################################################
#  CLUSTER 10                                                                  #
#  TYPE:  T/B BIPHENOTYPIC                                                     #
#  RNA MARKERS: CD7, CD3E, CD79A, IGHM, MS4A1                                  #
#  ADT MARKERS: CD19, CD21, IgD, CD20, IgM, CD2, CD5, CD3                      #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 10); display_top(dsb_de, 10)
# Update cell_types
cell_types <- update_cell_types(cell_types, 10, cell_type = "T/B biphenotypic")
# #
# # Since this is an unusual cell type, let's see whether these are mainly patient cells
# clus10_counts_by_group <- scler_pbmc@meta.data %>% filter(seurat_clusters == 10) %>% pull(group) %>% table()
# all_counts_by_group <- scler_pbmc@meta.data %>% pull(group) %>% table()
# clus10_prop_by_group <- clus10_counts_by_group / all_counts_by_group
# clus10_prop_by_group
# # Yes, this cluster includes a greater proportion of patient cells at all time
# # points than healthy cells.
################################################################################
#  CLUSTER 11                                                                  #
#  Source for pDC markers:                                                     #
#  https://www.beckman.com/resources/cell-types/blood-cells/leukocytes/        #
#  dendritic-cells/plasmacytoid-cells                                          #
#  TYPE: pDC                                                                   #
#  RNA MARKERS: IL3RA (CD123)                                                  #
#  ADT MARKERS: CD123                                                          #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 11); display_top(dsb_de, 11)
# # Look at IL3RA
# rna_de %>% filter(cluster == 11 & gene == "IL3RA")
# Update cell_types
cell_types <- update_cell_types(cell_types, 11, cell_type = "pDC")
################################################################################
#  CLUSTER 12                                                                  #
#  TYPE: NATURAL KILLER/T CELL                                                 #
#  RNA MARKERS: NCAM1                                                          #
#  ADT MARKERS: CD7, CD2, CD3, CD5                                             #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 12); display_top(dsb_de, 12)
# Update cell_types
cell_types <- update_cell_types(cell_types, 12, cell_type = "natural killer/T cell")
################################################################################
#  CLUSTER 13                                                                  #
#  TYPE: HSPC                                                                  #
#  RNA MARKERS: CD34, MYB                                                      #
#  ADT MARKERS:                                                                #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 13); display_top(dsb_de, 13)
# # CD34 and MYB are HSPC markers
# hspc_genes <- c("CD34", "MYB")
# # Violin plot
# VlnPlot(scler_pbmc, hspc_genes)
# Update cell_types
cell_types <- update_cell_types(cell_types, 13, cell_type = "HSPC")
################################################################################
#  CLUSTER 14                                                                  #
#  TYPE: PLATELET                                                              #
#  RNA MARKERS: PF4, PPBP, TUBB1, GP1BA, THBS1                                 #
#  ADT MARKERS: CD41                                                           #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 14); display_top(dsb_de, 14)
# # Define platelet markers
# platelet_genes <- c("PF4", "PPBP", "TUBB1", "GP1BA", "THBS1")
# # Visualize with violin plot
# VlnPlot(scler_pbmc, features = platelet_genes)
# Update cell_types
cell_types <- update_cell_types(cell_types, 14, cell_type = "platelet")
################################################################################
#  CLUSTER 15                                                                  #
#  TYPE: NATURAL KILLER                                                        #
#  RNA MARKERS: NCAM1                                                          #
#  ADT MARKERS: CD56, CD16                                                     #
################################################################################
# # Look at top markers for each layer
# display_top(rna_de, 15); display_top(dsb_de, 15)
# # Look at NCAM1
# rna_de %>% filter(cluster == 15 & gene == "NCAM1")
# Update cell_types
cell_types <- update_cell_types(cell_types, 15, cell_type = "natural killer")

# Add cell types to Seurat object
cell_type_vec <-
   setNames(as.character(cell_types$cell_type),
          as.character(cell_types$seurat_clusters))[as.character(scler_pbmc$seurat_clusters)]
names(cell_type_vec) <- NULL
scler_pbmc$cell_type <- cell_type_vec

# Save as H5
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/scler_pbmc_10"))
```

Ok, now we generate some manuscript figures. 

We start with heatmaps and dot plots showing the RNA and ADT markers for each cell type

```{r}
# Load Seurat objects with labeled cell types
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_10.h5seurat"))

# Set Idents to cell_type
Idents(scler_pbmc) <- "cell_type"

# FindAllMarkers by cell type
cell_type_markers_RNA <- FindAllMarkers(scler_pbmc, assay = "RNA")
cell_type_markers_ADT <- FindAllMarkers(scler_pbmc, assay = "DSB")

# Monocyte markers
# Lafyatis paper cites FCN1, S100A8, and S100A9 as monocyte markers
# https://acrjournals.onlinelibrary.wiley.com/doi/full/10.1002/art.41813?saml_referrer 
monocyte_markers_RNA <- c("FCN1", "S100A8", "S100A9")
monocyte_markers_ADT <- c("anti-human-CD14")

# B cell markers
# CD22: https://www.biocompare.com/Editorial-Articles/570578-A-Guide-to-B-Cell-Markers/
B_cell_markers_RNA <- c("CD79A", "MS4A1", "BANK1")
B_cell_markers_ADT <- c("anti-human-CD19")

# T cell markers
T_cell_markers_RNA <- c("CD3D", "CD3E")
CD4_T_cell_markers_RNA <- c("IL7R")
CD8_T_cell_markers_RNA <- c("CD8A", "CD8B")
T_cell_markers_ADT <- c("anti-human-CD3")
CD4_T_cell_markers_ADT <- c("anti-human-CD4")
CD8_T_cell_markers_ADT <- c("anti-human-CD8")

# natural killer/T markers
NK_T_markers_RNA <- c("STAT4", "RORA")

# NK markers
NK_markers_RNA <- c("GNLY", "NKG7", "KLRB1")
NK_markers_ADT <- c("anti-human-CD56", "anti-human-CD16")

# mDC markers
# https://www.biocompare.com/Editorial-Articles/572982-Dendritic-Cell-Markers/
mDC_markers_RNA <- c("CD1C", "CLEC10A")
mDC_markers_ADT <- c("anti-human-HLA-DR")

# pDC markers
# https://www.biocompare.com/Editorial-Articles/572982-Dendritic-Cell-Markers/
# https://en.wikipedia.org/wiki/LILRA4
pDC_markers_RNA <- c("IL3RA", "LILRA4", "CLEC4C")
pDC_markers_ADT <- c("anti-human-CD123")

# HSPC markers
# https://pmc.ncbi.nlm.nih.gov/articles/PMC9564379/
HSPC_markers_RNA <- c("CD34", "MYB", "KIT")
HSPC_markers_ADT <- c("anti-human-CD99")

# Platelet markers
platelet_markers_RNA <- c("PF4", "PPBP", "TUBB1")
platelet_markers_ADT <- c("anti-human-CD41")

# Put all RNA markers together
rna_markers <- c(monocyte_markers_RNA,
                 B_cell_markers_RNA,
                 T_cell_markers_RNA,
                 CD4_T_cell_markers_RNA,
                 CD8_T_cell_markers_RNA,
                 NK_T_markers_RNA,
                 NK_markers_RNA,
                 mDC_markers_RNA,
                 pDC_markers_RNA,
                 HSPC_markers_RNA,
                 platelet_markers_RNA)

# Put all ADT markers together
adt_markers <- c(monocyte_markers_ADT,
                 B_cell_markers_ADT,
                 T_cell_markers_ADT,
                 CD4_T_cell_markers_ADT,
                 CD8_T_cell_markers_ADT,
                 NK_markers_ADT,
                 mDC_markers_ADT,
                 pDC_markers_ADT,
                 HSPC_markers_ADT,
                 platelet_markers_ADT)

# Let's put our cell types in the same order as the markers
scler_pbmc$cell_type <- factor(scler_pbmc$cell_type,
                               levels = c("monocyte",
                                          "B cell",
                                          "T/B biphenotypic",
                                          "CD4+ T cell",
                                          "CD8+ T cell",
                                          "natural killer/T cell",
                                          "natural killer",
                                          "mDC",
                                          "pDC",
                                          "HSPC",
                                          "platelet"))

# Update idents to reflect re-ordering of cell_type variable
Idents(scler_pbmc) <- scler_pbmc$cell_type

# Heatmap showing RNA markers
rna_marker_heatmap <- 
    DoHeatmap(subset(scler_pbmc, downsample = 100),
              features = rna_markers,
              assay = "RNA",
              group.bar.height = 0,
              size = 5,
              lines.width = 3) +
    scale_fill_gradient(low = "grey95", high = "darkblue", na.value = "white") +
  #ggtitle("RNA markers") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15, color = "black"),
          plot.title = element_text(size = 25, face = "bold")) +
    guides(color = "none")
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/rna_marker_heatmap.png"), rna_marker_heatmap, width = 7, height = 9)

# Map full length antibody names to shorted names (e.g., anti-human-CD3 -> CD3)
label_map <- setNames(gsub("^anti-human-", "", adt_markers), adt_markers)
# Heatmap showing ADT markers
adt_marker_heatmap <- 
    DoHeatmap(subset(scler_pbmc, downsample = 100),
              features = adt_markers,
              assay = "DSB",
              group.bar.height = 0,
              size = 5,
              lines.width = 8,
              disp.min = -0.5,
              disp.max = 1) +
    scale_fill_gradient(low = "grey95", high = "darkred", na.value = "black") +
    scale_y_discrete(labels = label_map) +
  #ggtitle("ADT markers") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15, color = "black"),
          plot.title = element_text(size = 25, face = "bold")) +
    guides(color = "none")
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/adt_marker_heatmap.png"), adt_marker_heatmap, width = 7, height = 9)

# Dotplot showing RNA markers
rna_marker_dotplot <- 
  DotPlot(scler_pbmc,
          features = rna_markers,
          group.by = "cell_type",
          assay = "RNA") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          text = element_text(color = "black")) +
  labs(x = "Gene", y = "Cell type")
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/rna_marker_dotplot.png"), rna_marker_dotplot, width = 12, height = 8, bg = "white")

# Dotplot showing ADT markers
adt_marker_dotplot <- 
  DotPlot(scler_pbmc,
          features = adt_markers,
          group.by = "cell_type",
          assay = "DSB",
          cols = c("lightgrey", "darkred")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          text = element_text(color = "black")) +
  scale_x_discrete(labels = label_map) +
  labs(x = "Protein", y = "Cell type")
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/adt_marker_dotplot.png"), adt_marker_dotplot, width = 12, height = 8)

# Set levels for LIBRARY variable
scler_pbmc$LIBRARY <- factor(scler_pbmc$LIBRARY, levels = c("H", "515", "591", "594"))

# We also save the Seurat object to update the reordering of cell types
SaveH5Seurat(scler_pbmc, paste0(project_path, "data/h5seurat/scler_pbmc_11.h5seurat"))
```

Below we generate some UMAPs.

```{r}
# Load in Seurat object
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_11.h5seurat"))

# Set Idents of scler_pbmc to cell_type
Idents(scler_pbmc) <- scler_pbmc$cell_type

# UMAP for all jSSc and HC cells together
all_umap <- 
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T) +
  theme(legend.position = "none") +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2")
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/all_umap.png"), all_umap, height = 6, width = 5)

# UMAP split by health (jSSc vs. healthy)
all_umap_by_health <-
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T, split.by = "health")  +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2") +
  facet_wrap(~health, labeller = labeller(health = c(
    "healthy"           = "Healthy",
    "SSc" = "SSc"
  )), nrow = 1)
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/all_umap_by_health.png"), all_umap_by_health, width = 9, height = 4.5)

# UMAP split by group (jSSc timepoints + HC)
all_umap_by_group <-
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T, split.by = "group") +
  theme(legend.position = "none") +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2") +
  facet_wrap(~group, labeller = labeller(group = c(
    "healthy"           = "Healthy",
    "baseline" = "SSc baseline",
    "six"      = "SSc 6 mo.",
    "twelve"     = "SSc 12 mo.",
    "twen4"     = "SSc 24 mo."
  )), nrow = 1)
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/all_umap_by_group.png"), all_umap_by_group, width = 16, height = 4.5)

# UMAP split by library
all_umap_by_library <-
scler_pbmc %>% DimPlot(label = T, cols = brewer.pal(12, "Set3"), repel = T, split.by = "LIBRARY") +
  theme(legend.position = "none") +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2") +
  facet_wrap(~LIBRARY, labeller = labeller(LIBRARY = c(
    "H"           = "Healthy",
    "515" = "515",
    "591"      = "591",
    "594"     = "594"
  )), nrow = 1)
# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/all_umap_by_library.png"), all_umap_by_library, width = 14, height = 5.72)
```

Now we plot cell type proportions.

```{r}
# Turn cell type into a factor (ordered from most abundant to rarest)
scler_pbmc$cell_type <- factor(scler_pbmc$cell_type,
                               levels = c("monocyte",
                                          "B cell",
                                          "T/B biphenotypic",
                                          "CD4+ T cell",
                                          "CD8+ T cell",
                                          "natural killer/T cell",
                                          "natural killer",
                                          "mDC",
                                          "pDC",
                                          "HSPC",
                                          "platelet"))

# Add a blank level between HC and SSc baseline
scler_pbmc@meta.data$group_spaced <- factor(
  scler_pbmc@meta.data$group,
  levels = c(levels(scler_pbmc@meta.data$group)[1], "", 
             levels(scler_pbmc@meta.data$group)[2:5])
)

# Plot cell type proportions
prop_by_group <- ggplot(scler_pbmc@meta.data, aes(y = group_spaced, fill = cell_type)) +
    geom_bar(position = "fill", color = "black") +
    scale_fill_brewer(palette = "Set3", name = "Cell type") +
    scale_y_discrete(labels = c("Healthy", "", "SSc baseline", "SSc 6 mo.", "SSc 12 mo.", "SSc 24 mo."),
                     drop = FALSE) +
    coord_flip() +
    theme_bw() +
    labs(x = "Cell type proportion") +
    geom_hline(yintercept = 2, col = "black", linewidth = 1, linetype = "dashed") +
    theme(
        axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 18),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 18, face = "bold"),
        legend.position = "none"
    )

# Save as PNG
ggsave(paste0(project_path, "figures/2_pbmc_cell_type_annotation/prop_by_group.png"), prop_by_group, height = 7, width = 6)
```



