---
title: "3: jSSc ASCT monocyte subtype annotation"
output: html_notebook
---

This notebook follows 2_pbmc_cell_type_annotation.Rmd, in which we annotated all major subtypes. In this notebook, we annotate monocyte subtypes, since these cells are the focus of our analysis.

We begin by loading in useful packages and custom functions.

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater", "ggrepel", "tidyverse", "RColorBrewer", "pheatmap", "clusterProfiler", "enrichplot", "msigdbr")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))

# Function to update cell_types
update_cell_types <- function(df, cluster, cell_type = NA) {
  df[df$seurat_clusters == cluster, "cell_type"] <- cell_type
  return(df)
}
```

We load in the monocytes and cluster them. 

```{r}
# # scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_10.h5seurat"))
# # monocytes <- subset(scler_pbmc, cell_type == "monocyte")
# # 
# # monocytes <- sub_cluster(monocytes,
# #                          res = 0.8)
# # SaveH5Seurat(monocytes, paste0(project_path, "data/h5seurat/monocytes.h5seurat"))
```

Now we're going to annotate the monocytes.

```{r}
# monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes.h5seurat"))
# cell_types <- data.frame(seurat_clusters = levels(monocytes$seurat_clusters),
#                           cell_type = NA)
# 
# monocytes %>% DimPlot(reduction = "wnn.umap", label = T)
# 
# rna_markers <- FindAllMarkers(monocytes, assay = "RNA", only.pos = T)
# adt_markers <- FindAllMarkers(monocytes, assay = "DSB", only.pos = T)
# 
# monocytes$seurat_clusters %>% levels() %>% lapply(function(clus){
#   rna_markers %>% filter(cluster == clus) %>% head() %>% rbind((
#   adt_markers %>% filter(cluster == clus) %>% head()))
# }) %>% setNames(paste0("cluster_", (monocytes$seurat_clusters %>% levels())))
# 
# FeaturePlot(monocytes, "CD14", label = T, reduction = "wnn.umap")
# for (clus in c(0, 1, 2, 3, 4, 7, 9)) {
#   rna_markers %>% filter(cluster == clus & gene == "CD14") %>% rbind((
#   adt_markers %>% filter(cluster == clus & gene == "anti-human-CD14"))) %>% print()
#   cell_types <- update_cell_types(cell_types, clus, cell_type = "CD14+")
# }
# 
# FeaturePlot(monocytes, "FCGR3A", label = T, reduction = "wnn.umap")
# for (clus in c(5, 6)) {
#   rna_markers %>% filter(cluster == clus & gene == "FCGR3A") %>% rbind((
#   adt_markers %>% filter(cluster == clus & gene == "anti-human-CD16"))) %>% print()
#   cell_types <- update_cell_types(cell_types, clus, cell_type = "CD16+")
# }
# 
# # Now, we just need to figure out clusters 8, 10, and 11
# c(8, 10, 11) %>% lapply(function(clus){
#   rna_markers %>% filter(cluster == clus) %>% head() %>% rbind((
#   adt_markers %>% filter(cluster == clus) %>% head()))
# }) %>% setNames(paste0("cluster_", c(8, 10, 11)))
# 
# cell_types <- update_cell_types(cell_types, 8, "B-like")
# cell_types <- update_cell_types(cell_types, 10, "T-like")
# cell_types <- update_cell_types(cell_types, 11, "NK-like")
# 
# # Add cell types to Seurat object
# cell_type_vec <-
#    setNames(as.character(cell_types$cell_type),
#           as.character(cell_types$seurat_clusters))[as.character(monocytes$seurat_clusters)]
# names(cell_type_vec) <- NULL
# monocytes$cell_type <- cell_type_vec
# 
# 
# # Save annotated Seurat object
# SaveH5Seurat(monocytes, paste0(project_path, "data/h5seurat/monocytes_2.h5seurat"))
```

Ok, now we generate some manuscript figures. 

We start with heatmaps and dot plots showing the RNA and ADT markers for each monocyte subtype

```{r}
# Load in Seurat object
monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_2.h5seurat"))

# Set Idents to cell type
Idents(monocytes) <- "cell_type"

# FindAllMarkers by cell type
cell_type_markers_RNA <- FindAllMarkers(monocytes, assay = "RNA", only.pos = T)
cell_type_markers_ADT <- FindAllMarkers(monocytes, assay = "DSB", only.pos = T)

# CD14 markers
CD14_rna_markers <- c("CD14")
CD14_adt_markers <- c("anti-human-CD14")

# CD16 markers
CD16_rna_markers <- c("FCGR3A")
CD16_adt_markers <- c("anti-human-CD16")

# B-like markers
B_like_rna_markers <- c("CD79A", "MS4A1")
B_like_adt_markers <- c("anti-human-CD19", "anti-human-IgD")

# T-like markers
T_like_rna_markers <- c("CD3E", "CD3D")
T_like_adt_markers <- c("anti-human-CD3")

# NK-like markers
NK_like_rna_markers <- c("NKG7", "KLRD1")
NK_like_adt_markers <- c("anti-human-CD56")

# Put all RNA markers together
rna_cell_type_markers <- c(CD14_rna_markers,
                 CD16_rna_markers,
                 B_like_rna_markers,
                 T_like_rna_markers,
                 NK_like_rna_markers)

# Put all ADT markers together
adt_cell_type_markers <- c(CD14_adt_markers,
                 CD16_adt_markers,
                 B_like_adt_markers,
                 T_like_adt_markers,
                 NK_like_adt_markers)

# Order cell types to match order of markers
monocytes$cell_type <- factor(monocytes$cell_type,
                              levels = unique(monocytes$cell_type))

# Heatmap showing RNA markers
rna_marker_heatmap <-
    DoHeatmap(subset(monocytes, downsample = 60),
              features = rna_cell_type_markers,
              assay = "RNA",
              group.bar.height = 0,
              size = 5,
              lines.width = 3) +
    scale_fill_gradient(low = "grey95", high = "darkblue", na.value = "white") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15, color = "black"),
          plot.title = element_text(size = 25, face = "bold")) +
    guides(color = "none") #+
 # ggtitle("RNA markers")
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/monocytes_rna_marker_heatmap.png"), rna_marker_heatmap, width = 8, height = 6)

# Map full length antibody names to shorted names (e.g., anti-human-CD3 -> CD3)
label_map <- setNames(gsub("^anti-human-", "", adt_cell_type_markers), adt_cell_type_markers)
# Heatmap showing ADT markers
adt_marker_heatmap <-
    DoHeatmap(subset(monocytes, downsample = 60),
              features = adt_cell_type_markers,
              assay = "DSB",
              group.bar.height = 0,
              size = 5,
              lines.width = 8,
              disp.min = -0.5,
              disp.max = 1) +
    scale_fill_gradient(low = "grey95", high = "darkred", na.value = "black") +
      scale_y_discrete(labels = label_map) +
    #ggtitle("ADT markers") +
    theme(legend.position = "right",
          axis.text.y = element_text(size = 15, color = "black"),
          plot.title = element_text(size = 25, face = "bold")) +
  guides(color = "none")
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/monocytes_adt_marker_heatmap.png"), adt_marker_heatmap, width = 8, height = 6)

# Dotplot showing RNA markers
rna_marker_dotplot <-
  DotPlot(monocytes,
          features = rna_cell_type_markers,
          group.by = "cell_type",
          assay = "RNA") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          text = element_text(color = "black"))  +
  labs(x = "Gene", y = "Cell type")
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/monocytes_rna_marker_dotplot.png"), rna_marker_dotplot, width = 6, height = 6)

# Dotplot showing ADT markers
adt_marker_dotplot <-
  DotPlot(monocytes,
          features = adt_cell_type_markers,
          group.by = "cell_type",
          assay = "DSB",
          cols = c("lightgrey", "darkred")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          text = element_text(color = "black")) +
  scale_x_discrete(labels = label_map) +
  labs(x = "Protein", y = "Cell type")
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/monocytes_adt_marker_dotplot.png"), adt_marker_dotplot, width = 6, height = 6)
```

Now we generate some UMAPs.

```{r}
# All jSSc and HC monocytes
monocyte_dim_plot <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = rev(brewer.pal(8, "Set3"))) +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2") +
  theme(legend.position = "none",
        plot.title = element_blank())
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/jSSc_and_healthy_monocytes_umap.png"), height = 6, width = 5)

# By group (jSSc time points and HC)
monocyte_dim_plot_by_group <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = rev(brewer.pal(8, "Set3")), split.by = "group") +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2") +
  theme(plot.title = element_blank()) +
  facet_wrap(~group, labeller = labeller(group = c(
    "healthy"           = "HC",
    "baseline" = "SSc baseline",
    "six"      = "SSc 6 mo.",
    "twelve"     = "SSc 12 mo.",
    "twen4"     = "SSc 24 mo."
  )), nrow = 1)
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/jSSc_and_healthy_monocytes_by_group_umap.png"), height = 4, width = 11)

# By health (jSSc vs. HC)
monocyte_dim_plot_by_health <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = rev(brewer.pal(8, "Set3")), split.by = "health") +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2") +
  theme(plot.title = element_blank())
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/jSSc_and_healthy_monocytes_by_health_umap.png"), height = 5.79, width = 7.4)

# By library (3 patients and HC)
# We relabel H as HC, and turn into a factor for desired ordering
monocytes@meta.data <- monocytes@meta.data %>%
  mutate(
    LIBRARY = recode(LIBRARY, "H" = "HC"),
    LIBRARY = factor(LIBRARY, levels = c("HC", "515", "591", "594"))
  )
monocyte_dim_plot_by_library <-
  monocytes %>% DimPlot(reduction = "wnn.umap", label = T, group.by = "cell_type", cols = rev(brewer.pal(8, "Set3")), split.by = "LIBRARY") +
  labs(x = "WNN UMAP 1", y = "WNN UMAP 2") +
  theme(plot.title = element_blank())
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/jSSc_and_healthy_monocytes_by_library_umap.png"), height = 3, width = 10)
```

Now, we make a plot of monocyte subtype proportions.

```{r}
# Cell type proportions plot
# Add a blank level between HC and SSc baseline
monocytes@meta.data$group_spaced <- factor(
  monocytes@meta.data$group,
  levels = c(levels(monocytes@meta.data$group)[1], "", 
             levels(monocytes@meta.data$group)[2:5])
)

monocytes_prop_by_group <- ggplot(monocytes@meta.data, aes(y = group_spaced, fill = cell_type)) +
    geom_bar(position = "fill", color = "black") +
    scale_fill_manual(values = rev(brewer.pal(8, "Set3")), name = "Cell type") +
    scale_y_discrete(labels = c("HC", "", "SSc baseline", "SSc 6 mo.", "SSc 12 mo.", "SSc 24 mo."),
                     drop = FALSE) +
    coord_flip() +
    theme_bw() +
    labs(x = "Cell type proportion") +
    geom_hline(yintercept = 2, col = "black", linewidth = 1, linetype = "dashed") +
    #ggtitle(str_wrap("HC vs. jSSc monocyte subtype proportions over time since ASCT", width = 40)) +
    theme(axis.text.x = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 18),
        plot.title = element_text(size = 18, face = "bold"),
        legend.position = "none"
        #legend.text = element_text(size = 14)
    )
# Save as PNG
ggsave(paste0(project_path, "figures/3_monocyte_subtype_annotation/monocytes_prop_by_group.png"), monocytes_prop_by_group, height = 7, width = 6)
```

For biological annotation of monocyte sub-clusters:

```{r}
# Set idents
Idents(monocytes) <- "seurat_clusters"
# Get RNA markers
rna_markers <- FindAllMarkers(monocytes, assay = "RNA", only.pos = T)
# Only get signifiant markers
rna_markers_sig <- rna_markers %>% filter(avg_log2FC > 0.5 & p_val < 0.05)
# Split rna marker df by cluster
rna_marker_list <- split(rna_markers_sig, rna_markers_sig$cluster)
# Save as separate CSVs
lapply(names(rna_marker_list), function(clus){
  # Get cell type
  cell_type <- monocytes@meta.data %>% filter(seurat_clusters == clus) %>% pull(cell_type) %>% unique()
  # Cell type directory
  save_dir <- paste0(project_path, "data/annotation_deg_CSVs/monocyte_annotation/", cell_type)
  # Does directory for this cell type exist? If not create it
  if (!dir.exists(save_dir)) {dir.create(save_dir)}
  # Save CSV to cell type directory
  write_csv(rna_marker_list[[clus]], paste0(project_path, "data/annotation_deg_CSVs/monocyte_annotation/", cell_type, "/cluster_", clus, ".csv"))})
```


