---
title: "4: jSSc ASCT monocytes longitudinal analysis"
output: html_notebook
---

In this notebook, we run our longitudinal monocyte pseudobulk offset analysis. Adapted from Lee and Han (2024).

We begin by loading in useful packages and custom functions.

First, clear the environment and set the path where the project folder is saved.

```{r}
# Clear all objects from environment
rm(list = ls(all.names = TRUE))
# Free up memory
gc()
# Increase max vector memory
mem.maxVSize(200000)

project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"

results_dir <- paste0(project_path, "data/DEG_pathway_CSVs/")
```

Now load in useful packages and custom functions.

```{r}
# Load packages
packages <- c("Seurat", "dplyr", "tidyr", "SeuratDisk", "harmony", "readr",
              "ggplot2", "stringr", "SoupX", "dsb", "data.table", "Matrix",
              "scDblFinder", "scater", "ggrepel","tidyverse", "RColorBrewer", "pheatmap",
              "clusterProfiler", "enrichplot", "msigdbr", "patchwork", "grid", "gridExtra", "ggtext", "ggh4x", "gtable")
invisible(lapply(packages, require, character.only = TRUE))

# Source functions
function_files <- list.files(paste0(project_path, "functions"), pattern = "\\.R$", full.names = TRUE)
invisible(lapply(function_files, source))
```

We run the model on all the patient monocytes. 

```{r}
# Read in monocytes
monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_2.h5seurat"))

# Subset out patient monocytes
SSc_monocytes <- subset(monocytes, health == "SSc")

# Design
design_formula <- ~ subject + poly(timepoint)

# Fit offset pseudobulk model
pbo_monocytes <- pb_offset(
  seurat_obj = SSc_monocytes,
  sample_id_variable = "sample_id",
  design_formula = design_formula)

# Test for significant effect of time since ASCT
results <- test_de(pbo_monocytes$fit, `poly(timepoint)`) %>% 
  arrange(pval)

# Save all results
write_csv(results, paste0(results_dir, "all_DE_results.csv"))
```

Now we conduct pathway enrichment analysis based off of this tutorial:  https://biostatsquid.com/pathway-enrichment-analysis-tutorial-clusterprofiler/

Setup for GSEA is below:

```{r}
# Get hallmark, GO biological process, and GO cellular component pathways
hallmark_sets <- msigdbr(species = "human",
                         collection = "H")
bp_sets <- msigdbr(species = "human",
                   collection = "C5",
                   subcollection = "GO:BP")
cc_sets <- msigdbr(species = "human",
                   collection = "C5",
                   subcollection = "GO:CC")
# Store all sets together
all_sets <- rbind(hallmark_sets, bp_sets, cc_sets)

# TERM2GENE
term2gene <- all_sets %>%
  select(gs_name, gene_symbol)

# TERM2NAME (readable descriptions)
term2name <- all_sets %>%
  distinct(gs_name, gs_description)

# Background genes
DefaultAssay(SSc_monocytes) <- "RNA"
background_genes <- rownames(SSc_monocytes)
```

Let's run pathway analysis for each direction separately.
 
```{r}
# (1) PBMC

# Add direction (whether expression goes up or down over time) to results data frame
results_sig <- results %>% filter(pval < 0.05) %>% mutate(direction = case_when(
  lfc > 0 ~ 'UP',
  lfc < 0 ~ 'DOWN',
))

# Split the dataframe into a list of sub-dataframes: upregulated, downregulated genes
results_list <- split(results_sig, results_sig$direction)

# Run enricher R on each set of genes (up set and down set)
gsea <- lapply(names(results_list),
              function(x) enricher(
                  gene = results_list[[x]]$name,
                  TERM2GENE = term2gene,
                  TERM2NAME = term2name,
                  universe = background_genes,
                  pvalueCutoff = 1,
                  qvalueCutoff = 1,
                  minGSSize = 5,
                  maxGSSize = 500
              )) %>% setNames(names(results_list))

# View results
# DOWN
gsea$DOWN@result %>% head()
# UP
gsea$UP@result %>% head()
```

Now, we save
a) genes that go down
b) genes that go up
c) pathways that go down
d) pathways that go up

```{r}
# # Down genes
down_res <- results_sig %>% filter(lfc < 0)
write_csv(down_res, paste0(results_dir, "down_DEGs.csv"))

# # Up genes
up_res <- results_sig %>% filter(lfc > 0)
write_csv(up_res, paste0(results_dir, "up_DEGs.csv"))

# # Down pathways
down_pathways <- gsea$DOWN@result
write_csv(down_pathways, paste0(results_dir, "down_GSEA.csv"))

# # Up pathways
up_pathways <- gsea$UP@result
write_csv(up_pathways, paste0(results_dir, "up_GSEA.csv"))
```



Out of the top statistically significant pathways with increasing and decreasing activity, we select the most biologically relevant pathways and display them in bar charts below.

First, pathways with decreased activity over time since ASCT:

```{r}
# Data frame mapping pathway IDs to formal pathway names
down_pathways_to_show <- data.frame(full_name = c("HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                           "HALLMARK_INFLAMMATORY_RESPONSE",
                           "GOBP_EPITHELIAL_CELL_DIFFERENTIATION",
                           "HALLMARK_TGF_BETA_SIGNALING",
                           "HALLMARK_IL2_STAT5_SIGNALING",
                           "HALLMARK_APOPTOSIS"
                           ), plot_name = c("TNF-α signaling through NF-κB (Hallmark)",
                                            "Inflammatory response (Hallmark)",
                                            "Epithelial cell differentiation (GOBP)",
                                            "TGF-β signaling (Hallmark)",
                                            "IL2-STAT-5 signaling (Hallmark)",
                                            "Apoptosis (Hallmark)"))

# Create data frame including results for these pathways
gsea_down_df <- down_pathways_to_show %>%
  inner_join(
    gsea$DOWN@result %>% select(ID, p.adjust),
    by = c("full_name" = "ID")
  ) %>%
  filter(p.adjust < 0.05)

# Bar chart
gsea_down_plot <- ggplot(data = gsea_down_df, 
                         aes(x = -log10(p.adjust),
                             y = reorder(plot_name, -p.adjust),
                             fill = reorder(plot_name, -p.adjust))) +
  geom_col() +
  scale_fill_brewer(palette = "Blues") +
  theme_bw() +
  ylab("Gene set") +
  xlab(expression("-log"[10]*"p-value")) +
  ggtitle(str_wrap("Gene sets with decreased activity in jSSc monocytes over time since ASCT", 40)) +
  theme(
    text = element_text(face = "bold"),
    axis.text = element_text(color = "black", size = 13),
    axis.title = element_text(color = "black", size = 15),
    plot.title = element_text(color = "black", size = 15),
    legend.position = "none"
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 25))
# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/gsea_down_plot.png"), gsea_down_plot, width = 6.5, height = 5)
```

Now, pathways with increasing activity over time since ASCT.

```{r}
# Data frame mapping pathway IDs to formal names
up_pathways_to_show <- data.frame(full_name = c("GOBP_CELLULAR_RESPIRATION",
                         "GOBP_PROTON_MOTIVE_FORCE_DRIVEN_ATP_SYNTHESIS",
                         "GOBP_ELECTRON_TRANSPORT_CHAIN",
                         "GOCC_RESPIRATORY_CHAIN_COMPLEX",
                         "HALLMARK_OXIDATIVE_PHOSPHORYLATION"),
                         plot_name = c("Cellular respiration (GOBP)",
                                       "Proton motive force driven ATP synthesis (GOBP)",
                                       "Electron transport chain (GOBP)",
                                       "Respiratory chain complex (GOCC)",
                                       "Oxidative phosphorylation (Hallmark)"))

# Create data frame including results for these pathways
gsea_up_df <- up_pathways_to_show %>%
  inner_join(
    gsea$UP@result %>% select(ID, p.adjust),
    by = c("full_name" = "ID")
  ) %>%
  filter(p.adjust < 0.05)

# Bar chart
gsea_up_plot <- ggplot(data = gsea_up_df, 
                         aes(x = -log10(p.adjust),
                             y = reorder(plot_name, -p.adjust),
                             fill = reorder(plot_name, -p.adjust))) +
  geom_col() +
  scale_fill_brewer(palette = "Reds") +
  theme_bw() +
  ylab("Gene set") +
  xlab(expression("-log"[10]*"p-value")) +
  ggtitle(str_wrap("Gene sets with increased activity \nin jSSc monocytes over time since ASCT", 40)) +
  theme(
    text = element_text(face = "bold"),
    axis.text = element_text(color = "black", size = 13),
    axis.title = element_text(color = "black", size = 15),
    plot.title = element_text(color = "black", size = 15),
    legend.position = "none"
  ) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 20))
# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/gsea_up_plot.png"), gsea_up_plot, width = 6.5, height = 5)
```

Now, we make a tile plot showing the top 5 most statistically significant longitudinal DEGs for our top decreasing pathways.

```{r}
# Get all gene sets
gene_sets_down <- gsea$DOWN@geneSets
# Select top pathways
favorite_down_pathways <- down_pathways_to_show$full_name
fav_gene_sets_down <- gene_sets_down[favorite_down_pathways]

# Make a named vector of p-values for quick lookup
pval_lookup_down <- results %>% filter(lfc < 0) %>% pull(pval)
names(pval_lookup_down) <- results %>% filter(lfc < 0) %>% pull(name)

# Order each pathway's genes by p-value
fav_gene_sets_down_ordered <- lapply(fav_gene_sets_down, function(genes) {
  # Keep only genes that exist in results
  genes_in_results <- intersect(genes, names(pval_lookup_down))
  
  # Order by p-value
  genes_in_results[order(pval_lookup_down[genes_in_results])]
})

# Get top 5 genes for each pathway and take intersection
topgenes_down <- lapply(fav_gene_sets_down_ordered, function(x){head(x, 5)}) %>% unlist() %>% setNames(NULL) %>% unique()

# Matrix indicating which genes are in which pathway
pathway_by_gene_mat_down <- sapply(topgenes_down, function(gene) {
  sapply(fav_gene_sets_down_ordered, function(x) gene %in% x)
})

# Turn into data frame
heat_df_down <- pathway_by_gene_mat_down %>%
  as.data.frame() %>%
  rownames_to_column("pathway") %>%
  pivot_longer(-pathway, names_to = "gene", values_to = "present")

# Add gene to DF
heat_df_down$gene <- factor(heat_df_down$gene, levels = topgenes_down)

# Left join with gsea_down_df (formal pathway names for plotting)
heat_df_down <-
heat_df_down %>%
    left_join(
        gsea_down_df %>% select(full_name, plot_name),
        by = c("pathway" = "full_name")
    )

# We use same palette as prior pathway plot
down_gb <- ggplot_build(gsea_down_plot)
down_fill_mapping <- 
data.frame(
  full_name = down_gb$plot$data$full_name,
  fill = unique(down_gb$data[[1]]$fill)
)


# Create new variable with fill color
heat_df_down <- heat_df_down %>%
  left_join(down_fill_mapping, by = c("pathway" = "full_name")) %>%
  mutate(fill_color = ifelse(present, fill, "white"))


# Genes to bold (in 3 + pathways)
pathway_counts <- levels(heat_df_down$gene) %>% lapply(function(GENE){heat_df_down %>% filter(gene == GENE) %>% pull(present) %>% sum()}) %>% setNames(levels(heat_df_down$gene)) %>% unlist()
genes_to_bold <- pathway_counts[pathway_counts >= 3] %>% names()

# Named vector: gene -> styled label
gene_labels <- heat_df_down %>%
    distinct(gene) %>%
    mutate(label = ifelse(gene %in% genes_to_bold,
                          paste0("<b>", gene, "</b>"),
                          as.character(gene))) %>%
    tibble::deframe()

# Set levels of pathway
heat_df_down$plot_name <- factor(
  heat_df_down$plot_name,
  levels = down_gb$layout$panel_params[[1]]$y$breaks
)

# Plot it
gsea_plot_with_genes_down <- 
  ggplot(heat_df_down, aes(x = gene, y = plot_name, fill = fill_color)) +
    geom_tile(color = "black") +
    theme_minimal() +
  scale_fill_identity() +
    theme(
      panel.grid = element_blank(),
      legend.position = "none",
    axis.text = element_text(color = "black", size = 13),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(color = "black", size = 15, face = "bold"),
    plot.title = element_text(color = "black", size = 15, face = "bold"),
    ) +
    ylab("Gene Set") +
    xlab("Gene") +
    ggtitle(str_wrap("Top DEGs contributing to pathways with decreased activity in jSSc monocytes over time since ASCT", width = 40)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 25)) +
  scale_x_discrete(labels = gene_labels) +
  theme(axis.text.x = element_markdown(angle = 45, hjust = 1, size = 11))

ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/gsea_down_plot_with_genes.png"), gsea_plot_with_genes_down, width = 7, height = 5, bg = "white")
```

We do the same thing for increasing genes (though this is not the focus of our paper; will likely include in supplement)

```{r}
# Get all gene sets
gene_sets_up <- gsea$UP@geneSets
# Select top 5 pathways
favorite_up_pathways <- up_pathways_to_show$full_name
fav_gene_sets_up <- gene_sets_up[favorite_up_pathways]

# Make a named vector of p-values for quick lookup
pval_lookup_up <- results %>% filter(lfc > 0) %>% pull(pval)
names(pval_lookup_up) <- results %>% filter(lfc > 0) %>% pull(name)

# Order each pathway's genes by p-value
fav_gene_sets_up_ordered <- lapply(fav_gene_sets_up, function(genes) {
  # Keep only genes that exist in results
  genes_in_results <- intersect(genes, names(pval_lookup_up))
  
  # Order by p-value
  genes_in_results[order(pval_lookup_up[genes_in_results])]
})

# Get top 5 genes for each pathway and take intersection
topgenes_up <- lapply(fav_gene_sets_up_ordered, function(x){head(x, 5)}) %>% unlist() %>% setNames(NULL) %>% unique()

# Matrix indicating which genes are in which pathway
pathway_by_gene_mat_up <- sapply(topgenes_up, function(gene) {
  sapply(fav_gene_sets_up_ordered, function(x) gene %in% x)
})

# Turn into data frame
heat_df_up <- pathway_by_gene_mat_up %>%
  as.data.frame() %>%
  rownames_to_column("pathway") %>%
  pivot_longer(-pathway, names_to = "gene", values_to = "present")

# Add gene to DF
heat_df_up$gene <- factor(heat_df_up$gene, levels = topgenes_up)

# Left join with gsea_up_df (formal pathway names for plotting)
heat_df_up <-
heat_df_up %>%
    left_join(
        gsea_up_df %>% select(full_name, plot_name),
        by = c("pathway" = "full_name")
    )

# We use same palette as prior pathway plot
up_gb <- ggplot_build(gsea_up_plot)
up_fill_mapping <- 
data.frame(
  full_name = up_gb$plot$data$full_name,
  fill = unique(up_gb$data[[1]]$fill)
)

# Create new variable with fill color
heat_df_up <- heat_df_up %>%
  left_join(up_fill_mapping, by = c("pathway" = "full_name")) %>%
  mutate(fill_color = ifelse(present, fill, "white"))


# Genes to bold (in 3 + pathways)
pathway_counts <- levels(heat_df_up$gene) %>% lapply(function(GENE){heat_df_up %>% filter(gene == GENE) %>% pull(present) %>% sum()}) %>% setNames(levels(heat_df_up$gene)) %>% unlist()
genes_to_bold <- pathway_counts[pathway_counts >= 3] %>% names()

# Set levels of pathway
heat_df_up$plot_name <- factor(
  heat_df_up$plot_name,
  levels = up_gb$layout$panel_params[[1]]$y$breaks
)

# Named vector: gene -> styled label
gene_labels <- heat_df_up %>%
    distinct(gene) %>%
    mutate(label = ifelse(gene %in% genes_to_bold,
                          paste0("<b>", gene, "</b>"),
                          as.character(gene))) %>%
    tibble::deframe()

# Plot it
gsea_plot_with_genes_up <- 
  ggplot(heat_df_up, aes(x = gene, y = plot_name, fill = fill_color)) +
    geom_tile(color = "black") +
    theme_minimal() +
  scale_fill_identity() +
    theme(
      panel.grid = element_blank(),
      legend.position = "none",
      #text = element_text(face = "bold"),
    axis.text = element_text(color = "black", size = 13),
    axis.text.y = element_text(face = "bold"),
    axis.title = element_text(color = "black", size = 15, face = "bold"),
    plot.title = element_text(color = "black", size = 15, face = "bold")
    ) +
    ylab("Gene Set") +
    xlab("Gene") +
    ggtitle(str_wrap("Top DEGs contributing to pathways with decreased activity in jSSc monocytes over time since ASCT", width = 40)) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 25)) +
  scale_x_discrete(labels = gene_labels) +
  theme(axis.text.x = element_markdown(angle = 45, hjust = 1, size = 11))

ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/gsea_up_plot_with_genes.png"), gsea_plot_with_genes_up, width = 7, height = 5, bg = "white")
```

Now we will produce longitudinal plots for the genes shown in the tile plots.

We start with decreasing genes.

```{r}
# Plot with custom function
gene_plots_down <- plot_pb_offset(pbo_monocytes, topgenes_down, groupBy = "subject")

# Arrange plots
gene_plots_down_clean <- lapply(seq_along(gene_plots_down), function(i) {
  p <- gene_plots_down[[i]]
  nrow <- 7; ncol <- 3
  row <- (i - 1) %/% ncol
  col <- (i - 1) %% ncol
  p <- p + theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.title = element_text(size = 30)
  )
  p
})
# Customize
gene_plots_down_combined <- wrap_plots(gene_plots_down_clean, nrow = 7, ncol = 3) +
  plot_layout(guides = "collect") +
  plot_annotation(
    caption = "Months since transplant"
  ) &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 30, face = "plain"),   # fix bold/oversized
    legend.title = element_text(size = 30, face = "plain"),
    legend.key.size = unit(0.8, "lines"),
    plot.caption = element_text(hjust = 0.5, size = 30),
    plot.margin = margin(50, 10, 10, 30, "pt") 
  )
# Customize further
gene_plots_down_combined <- wrap_elements(full = gene_plots_down_combined) +
  labs(tag = "Counts per million (log2 scale)") +
  theme(
    plot.tag = element_text(size = 30, angle = 90, hjust = 0.5, vjust = -1),
    plot.tag.position = "left"
  ) + ggtitle("Downward longitudinal trends for key DEGs \nin jSSc monocytes over time since ASCT") + theme(plot.title = element_text(size = 30, face = "bold"))
# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/gene_plots_down_combined.png"), 
       gene_plots_down_combined, width = 14, height = 24, bg = "white")
```

Now for increasing genes/pathways.

```{r}
# Plot with custom function
gene_plots_up <- plot_pb_offset(pbo_monocytes, topgenes_up, groupBy = "subject")

# Arrange plots
gene_plots_up_clean <- lapply(seq_along(gene_plots_up), function(i) {
  p <- gene_plots_up[[i]]
  nrow <- 4; ncol <- 3
  row <- (i - 1) %/% ncol
  col <- (i - 1) %% ncol
  p <- p + theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.title = element_text(size = 30)
  )
  p
})
# Customize
gene_plots_up_combined <- wrap_plots(gene_plots_up_clean, nrow = 4, ncol = 3) +
  plot_layout(guides = "collect") +
  plot_annotation(
    caption = "Months since transplant"
  ) &
  theme(
    legend.position = "right",
    legend.text = element_text(size = 30, face = "plain"),   # fix bold/oversized
    legend.title = element_text(size = 30, face = "plain"),
    legend.key.size = unit(0.8, "lines"),
    plot.caption = element_text(hjust = 0.5, size = 30),
    plot.margin = margin(50, 10, 10, 30, "pt") 
  )
# Customize further
gene_plots_up_combined <- wrap_elements(full = gene_plots_up_combined) +
  labs(tag = "Counts per million (log2 scale)") +
  theme(
    plot.tag = element_text(size = 30, angle = 90, hjust = 0.5, vjust = -1),
    plot.tag.position = "left"
  ) + ggtitle("Upward longitudinal trends for key DEGs \nin jSSc monocytes over time since ASCT") + theme(plot.title = element_text(size = 30, face = "bold"))
# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/gene_plots_up_combined.png"), 
       gene_plots_up_combined, width = 14, height = 18, bg = "white")
```

Now, we make a reduced version of the downward DEGs to be included in one of the main figures.

```{r}
# Select group of decreasing DEGs to show
six_down_degs <- c("SERPINE1", "FOSL1", "SMAD3", "TAGLN2", "CDKN1C", "LYN")

# Plot with custom functions
gene_plots_down_reduced <- plot_pb_offset(pbo_monocytes, six_down_degs, groupBy = "subject")

# Clean individual plots
gene_plots_down_reduced_clean <- lapply(seq_along(gene_plots_down_reduced), function(i) {
  gene_plots_down_reduced[[i]] + theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.title = element_text(size = 30),
    plot.margin = margin(20, 10, 5, 30, "pt")
  )
})

# Caption as blank plot.
caption_plot <- ggplot() +
  annotate("text", x = 0.5, y = 0.5, label = "Months since transplant", size = 6.5) +
  theme_void() +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5, 50, 5, 50, "pt"))

# Stack plots + caption, collect legend, add title via plot_annotation
# Title goes here so patchwork reserves space for it properly above the plots
inner_with_caption <- wrap_plots(
    c(gene_plots_down_reduced_clean, list(caption_plot)),
    nrow = 7, ncol = 1,
    heights = c(rep(1, 6), 0.15)
  ) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = str_wrap("Downward longitudinal trends for key DEGs in jSSc monocytes over time since ASCT", width = 30),
    theme = theme(
      plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
      plot.margin = margin(10, 10, 10, 10, "pt")
    )
  ) &
  theme(
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.text = element_text(size = 20, face = "plain"),
    legend.title = element_text(size = 20, face = "plain"),
    legend.key.size = unit(0.8, "lines"),
    legend.box.background = element_rect(color = "black")
  )

# Wrap and add y-axis tag only
gene_plots_down_reduced_combined <- wrap_elements(full = inner_with_caption) +
  labs(tag = "Counts per million (log2 scale)") +
  theme(
    plot.tag = element_text(size = 30, angle = 90, hjust = 0.5, vjust = -1),
    plot.tag.position = "left",
    plot.margin = margin(10, 20, 10, 10, "pt")
  )

ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/gene_plots_down_reduced_combined.png"), 
       gene_plots_down_reduced_combined, width = 5, height = 20, bg = "white")
```

Now, we'll make some volcano plots showing DEGs that go up and down over time since ASCT. We highlight the same pathway-involved genes we used above, as well as the top 10 genes in each direction.

```{r}
# Top 10 genes that go up
top10_up <- results %>% arrange(pval) %>% filter(lfc > 0.5) %>% pull(name) %>% head(10)
# Key upward pathway genes
pathway_genes_up <- heat_df_up$gene %>% levels()
# Take union
up_highlight <- union(top10_up, pathway_genes_up)
# We don't want to show genes without a name
up_highlight <- up_highlight[!grepl("ENSG", up_highlight)]

# Top 10 genes that go down
top10_down <- results %>% arrange(pval) %>% filter(lfc < -0.5) %>% pull(name) %>% head(10)
# Key downward pathway genes
pathway_genes_down <- heat_df_down$gene %>% levels()
# Take union
down_highlight <- union(top10_down, pathway_genes_down)

# Create new data frame for plotting (mutation of results from pb offset)
volc_results <- results %>%
  mutate(`Trend following ASCT` = case_when(
    lfc > 0.5 & pval < 0.05 ~ "Up",
    lfc < -0.5 & pval < 0.05 ~ "Down",
    TRUE ~ "Not significant"
  )) %>%
  mutate(delabel = case_when(
    name %in% c(up_highlight, down_highlight) ~ name,
    TRUE ~ NA
  ))

# Put significance levels in proper order
volc_results$`Trend following ASCT` <- factor(volc_results$`Trend following ASCT`, levels = c("Up", "Down", "Not significant"))

# Volcano plot
jSSc_longitudinal_volcano <- ggplot(data = volc_results, aes(x = lfc, y = -log10(pval), col = `Trend following ASCT`, label = delabel)) +
  geom_vline(xintercept = c(-0.5, 0.5), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
  geom_point(size = 2) +
  theme_classic(base_size = 20) +
  theme(axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), color = 'black'),
        axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), color = 'black'),
        plot.title = element_text(hjust = 0.5),
        plot.title.position = "plot",
        text = element_text(face = "bold", color = "black", size = 25),
        legend.position = "inside",
        legend.position.inside = c(0.2, 0.2),
legend.background = element_rect(fill = alpha("white", 0.4)),
legend.box.background = element_rect(color = "black")) +
  scale_color_manual(values = c("red3", "#00AFBB", "grey")) +
  labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) + 
  scale_x_continuous(breaks = seq(-10, 10, 2)) +
  ggtitle(str_wrap("Differentially expressed genes over time since ASCT in jSSc monocytes", width = 50)) +
  geom_label_repel(#max.overlaps = Inf,
                  # size = rel(3),
                  box.padding = 0.3,
                  force = 4,
                  show.legend = F,
                  na.rm = T,
                  size = 6.5)

# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/jSSc_longitudinal_volcano.png"), jSSc_longitudinal_volcano, width = 12, height = 8)
```

We also make a volcano plot comparing jSSc and HC.

```{r}
# Make sure default assay is set to RNA
DefaultAssay(monocytes) <- "RNA"

# Run pb offset for baseline and healthy, where we are testing group membership (two-way comparison) rather than time coefficient.
SSc_v_h_PB <- pb_offset(subset(monocytes, group %in% c("baseline", "healthy")), "sample_id", ~health)
# Test DE
SSc_v_h_PB_results <- SSc_v_h_PB$fit %>% test_de("healthSSc") %>% arrange(pval)

# We save the top 50 upregulated DEGs and top 50 downregulated DEGs as CSV
top50_up_results <- SSc_v_h_PB_results %>% filter(lfc > 0.5) %>% head(50)
top50_down_results <- SSc_v_h_PB_results %>% filter(lfc < 0.5) %>% head(50)
# Write CSV
write_csv(top50_up_results, paste0(project_path, "data/DEG_pathway_CSVs/SSc_v_healthy/top50_SSc_v_healthy_monocytes_UP.csv"))
write_csv(top50_down_results, paste0(project_path, "data/DEG_pathway_CSVs/SSc_v_healthy/top50_SSc_v_healthy_monocytes_DOWN.csv"))

# Genes to highlight (biologically interesting)
ssc_v_h_up_highlight <- c("HLA-DRB5", "FOLR3", "THBS1", "TMEM176B", "EREG", "SNHG5", "PLCB1", "TEX14", "LYZ", "KIF9", "PLAUR", "FOS", "SLC2A3", "TMEM170B", "CCNL1", "KLF6", "BRK1")
ssc_v_h_down_highlight <- c("LINC03078", "ANK2", "HFM1", "RETN", "SLCO5A1", "TPTEP1", "CDKN1C", "PIGK", "RHOC", "POLR2J3", "TALAM1", "LILRB1", "HSPB1", "CTSL", "RRAS", "RPS26", "WARS1", "CD81", "RPL41","CRIP1")

# Create new data frame for plotting (mutation of results from pb offset)
SSc_v_h_PB_volc_results <- SSc_v_h_PB_results %>%
  mutate(`SSc vs. HC` = case_when(
    lfc > 0.5 & pval < 0.05 ~ "Up",
    lfc < -0.5 & pval < 0.05 ~ "Down",
    TRUE ~ "Not significant"
  )) %>%
  mutate(delabel = case_when(
    name %in% c(ssc_v_h_up_highlight, ssc_v_h_down_highlight) ~ name,
    TRUE ~ NA
  ))

# Put significance levels in proper order
SSc_v_h_PB_volc_results$`SSc vs. HC` <- factor(SSc_v_h_PB_volc_results$`SSc vs. HC`, levels = c("Up", "Down", "Not significant"))

# Volcano plot
SSc_v_h_volcano <- ggplot(data = SSc_v_h_PB_volc_results, aes(x = lfc, y = -log10(pval), col = `SSc vs. HC`, label = delabel)) +
  geom_vline(xintercept = c(-0.5, 0.5), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') +
  geom_point(size = 2) +
  theme_classic(base_size = 20) +
  theme(axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
        axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
        plot.title = element_text(hjust = 0.5, size = rel(1.1)),
        plot.title.position = "plot",
        text = element_text(face = "bold", color = "black")) +
  scale_color_manual(values = c( "red3", "#00AFBB", "grey")) +
  labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) +
  scale_x_continuous(breaks = seq(-10, 10, 2)) +
  ggtitle(str_wrap("Differentially expressed genes in SSc monocytes (before ASCT) vs. healthy monocytes (pseudobulk level)", width = 75)) +
  geom_label_repel(max.overlaps = Inf,
                  size = 6.5,
                  box.padding = 0.3,
                  force = 4,
                  show.legend = F,
                  na.rm = T)
# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/SSc_v_healthy_pb_volcano.png"), width = 12, height = 6)
```




Now, we make a dotplot showing the expression of some key downward DEGs. 

```{r}
# Create dot plot
down_deg_dot <-
DotPlot(monocytes,
        features = six_down_degs,
        group.by = "group",
        dot.scale = 30)

# Replace the default geom_point layer
down_deg_dot$layers[[1]] <-
  geom_point(
    aes(size = pct.exp, fill = avg.exp.scaled),
    shape = 21,
    color = "black",
    stroke = 0.6
  )

down_deg_dot <-
  down_deg_dot +
    coord_flip() +
    scale_fill_gradient2(low = "steelblue", mid = "white", high = "darkred", midpoint = 0) +
    ggtitle(str_wrap("Expression of select jSSc monocyte DEGs over time since ASCT", width = 70)) +
    ylab("Timepoint") +
    xlab("Gene") +
    theme(plot.caption = element_text(size = 9, hjust = 0))

# Save image
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/down_deg_dot.png"), down_deg_dot, width = 10, height = 6, bg = "white")
```

We assign a biological annotation to each cluster

```{r}
monocytes@meta.data <- monocytes@meta.data %>%
  mutate(
    bio_annotation = case_when(
      seurat_clusters == 0 ~ "CD14+ CCL20+ TK1+",
      seurat_clusters == 1 ~ "CD14+ PDE3A+ LY6G5B+",
      seurat_clusters == 2 ~ "CD14+ LAMP5+ MMP14+ MUC12+",
      seurat_clusters == 3 ~ "CD14+ CLEC5A+ CXCR2+ TNFAIP6+",
      seurat_clusters == 4 ~ "CD14+ ST7-AS1+ SINGLEC1+ ZNF219+",
      seurat_clusters == 5 ~ "CD16+ CDKN1C+ LYPD2+ MERTK+",
      seurat_clusters == 6 ~ "CD16+ LY6E+ FCGR3A+ IFITM1+",
      seurat_clusters == 7 ~ "CD14+ CCL2+ CXCL10+ IFI44L+",
      seurat_clusters == 9 ~ "CD14+ MARCO+ CCL2+ CLDN5+",
      TRUE ~ cell_type
    )
  )

# Break up into multiple lines for plotting
monocytes$bio_annotation_wrapped <- str_wrap(monocytes$bio_annotation, width = 15)


# Custom color palette
my_pal <- c(
  "#4E79A7", 
  "#F28E2B",  
  "#59A14F",  
  "#E15759", 
  "#76B7B2",  
  "#EDC948",  
  "#B07AA1",  
  "#FF9DA7",  
  "#6A5ACD",  
  "#00BCD4", 
  "#499894",  
  "#F06292"   
)

# Build UMAP
monocyte_bio_umap <-
  monocytes %>%
  DimPlot(
    reduction  = "wnn.umap",
    label      = TRUE,
    group.by   = "bio_annotation_wrapped",
    cols       = my_pal,
    pt.size    = 0.6,
    label.box  = TRUE,
    repel      = TRUE,
    combine    = FALSE
  ) %>%
  pluck(1) +
  ggtitle("jSSc and healthy monocyte sub-clusters with biological annotations") +
  theme(plot.title = element_text(size = 18))

# Find and style the label layer
label_idx <- which(sapply(
  monocyte_bio_umap$layers,
  function(l) inherits(l$geom, "GeomLabelRepel")
))

monocyte_bio_umap$layers[[label_idx]]$aes_params$fill      <- "white"
monocyte_bio_umap$layers[[label_idx]]$aes_params$alpha     <- 0.75
monocyte_bio_umap$layers[[label_idx]]$aes_params$fontface  <- "bold"
monocyte_bio_umap$layers[[label_idx]]$aes_params$size      <- 5
monocyte_bio_umap$layers[[label_idx]]$aes_params$linewidth <- 0.4

monocyte_bio_umap

# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/bio_annotation_umap.png"),
       monocyte_bio_umap, width = 10, height = 8)
```


Now we make a dotplot showing these markers:

```{r}
# Markers to show
bio_markers <- c("CCL20", "TK1",
                 "PDE3A", "LY6G5B",
                 "LAMP5", "MMP14", "MUC12",
                 "CLEC5A", "CXCR2", "TNFAIP6",
                 "ST7-AS1", "SIGLEC1", "ZNF219",
                 "CDKN1C", "LYPD2", "MERTK",
                 "LY6E", "FCGR3A", "IFITM1",
                 "CCL2", "CXCL10", "IFI44L",
                 "MARCO", "CLDN5")
mono_subset  <- monocytes %>% subset(!grepl("like", cell_type)) %>% subset(downsample = 60)

dp <- DotPlot(mono_subset, bio_markers, group.by = "bio_annotation_wrapped")
dp_data <- dp$data
dp_data$id <- factor(dp_data$id, levels = str_wrap(c("CD14+ CCL20+ TK1+",
"CD14+ PDE3A+ LY6G5B+",
"CD14+ LAMP5+ MMP14+ MUC12+",
"CD14+ CLEC5A+ CXCR2+ TNFAIP6+",
"CD14+ ST7-AS1+ SINGLEC1+ ZNF219+",
"CD16+ CDKN1C+ LYPD2+ MERTK+",
"CD16+ LY6E+ FCGR3A+ IFITM1+",
"CD14+ CCL2+ CXCL10+ IFI44L+",
"CD14+ MARCO+ CCL2+ CLDN5+"), width = 15))

ggplot(dp_data, aes(y = features.plot, x = id, fill = avg.exp.scaled)) +
  geom_tile()
```









We find cluster 9 (subgroup of CD14+ monocytes) particularly interesting, as many of are key decreasing genes are expressed in this cluster. We make some plots to illustrate this.

```{r}
clus9_umap_by_group <-
DimPlot(monocytes,
        reduction = "wnn.umap",
        label = TRUE,
        split.by = "group",
        cells.highlight = Cells(subset(monocytes, seurat_clusters == 9))) +
  scale_color_manual(
    values = c("Unselected" = "grey80", "Group_1" = "red"),
    breaks = "Group_1",
    labels = "Cluster 9"
  ) +
  labs(color = NULL) +
  ggtitle("Monocyte cluster 9 in healthy controls and jSSc patients over time since ASCT") +
  theme(plot.title.position = "plot",
        plot.title = element_text(size = 14)) +
  facet_wrap(~group, labeller = as_labeller(c(
    "healthy" = "Healthy",
    "baseline" = "Baseline",
    "six" = "6 mo.",
    "twelve" = "12 mo.",
    "twen4" = "24 mo."
  )), nrow = 1)
# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/clus9_umap_by_group.png"), width = 8, height = 4)
```

We use FindMarkers to determine which genes are upregulated in cluster 9 compared to other clusters. Then, we identify genes which are upregulated in cluster 9 and also involved in one of our gene sets of interest (decreasing pathways, which makes sense since cluster 9 disappears over time).

```{r}
# # Set Idents to clusters
# Idents(monocytes) <- "seurat_clusters"
# 
# # Find DEGs between clus 9 and other clusters
# clus9_markers <- FindMarkers(monocytes, ident.1 = 9, assay = "RNA")
# 
# # Get signficiant markers
# sig_clus_9_markers <- clus9_markers %>% filter(p_val_adj < 0.05 & avg_log2FC > 2) %>% rownames()
# 
# # Now, we see which markers appear in our pathways of interest.
# cluster_9_markers_per_pathway <- lapply(fav_gene_sets_down, function(gs){gs[gs %in% sig_clus_9_markers]})
# 
# # Subset out cluster 9 markers in our pathways
# clus9_markers_in_pathway <- clus9_markers[rownames(clus9_markers) %in% unique(unlist(cluster_9_markers_per_pathway)), ]
# 
# # We will label with top pathway
# clus9_markers_in_pathway$pathway <- NA
# 
# # Create gene column
# clus9_markers_in_pathway$gene <- rownames(clus9_markers_in_pathway)
# 
# # Add pathway (each gene is assigned to the most significant pathway it appears in)
# for (pathway in names(cluster_9_markers_per_pathway)) {
#   for (gene in clus9_markers_in_pathway$gene) {
#     if (clus9_markers_in_pathway[gene, ]$pathway %>% is.na) {
#       if (gene %in% cluster_9_markers_per_pathway[[pathway]]) {
#         clus9_markers_in_pathway[gene, ]$pathway <- pathway
#       }
#     }
#   }
# }
# 
# # Add pathway names for plotting
# clus9_markers_in_pathway <-
# clus9_markers_in_pathway %>%
#     left_join(
#         down_pathways_to_show %>% select(full_name, plot_name),
#         by = c("pathway" = "full_name")
#     )
# 
# # Group by pathway
# clus9_markers_in_pathway <- clus9_markers_in_pathway %>%
#   mutate(plot_name = factor(plot_name, levels = down_gb$layout$panel_params[[1]]$y$breaks)) %>%
#   arrange(plot_name)
# 
# # We use the dotplot function to get scaled averages for our heatmap
# dp <- DotPlot(monocytes, clus9_markers_in_pathway$gene, group.by = "seurat_clusters")
# # Extract data
# dp_data <- dp$data
# # Store levels of gene variable
# lvls <- levels(dp_data$features.plot)
# # Join
# dp_data <- dp_data %>%
#   left_join(
#     clus9_markers_in_pathway %>%
#       select(gene, plot_name),
#     by = c("features.plot" = "gene")
#   ) %>%
#   rename(pathway = plot_name) %>%
#   mutate(features.plot = factor(features.plot, levels = lvls))

# Build alternating fill vector matching your pathway factor levels
strip_fills <- rep(c("grey90", "white"),
                   length.out = length(levels(forcats::fct_rev(dp_data$pathway))))

# Heatmap
clus9_heatmap <-
    ggplot(dp_data, aes(y = features.plot, x = id, fill = avg.exp.scaled)) +
    geom_tile() +
    scale_fill_distiller(palette = "RdYlBu") +
    ggh4x::facet_grid2(                                                         # CHANGED
        forcats::fct_rev(pathway) ~ ., scales = "free_y", space = "free_y",
        switch   = "y",
        labeller = label_wrap_gen(width = 15),                                  # ADDED
        strip    = strip_themed(                                                 # ADDED
            background_y = elem_list_rect(fill = strip_fills, color = NA)       # ADDED
        )                                                                        # ADDED
    ) +
    labs(fill = str_wrap("Scaled average expression", 10),
         x = "Monocyte sub-cluster",
         y = "",
         title = str_wrap("Scaled average expression of key SSc pathway genes in jSSc and healthy monocytes by cluster", 65)) +
    theme(
        axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black", face = "bold"),
        legend.title = element_text(color = "black", face = "bold"),
        legend.text = element_text(color = "black", face = "bold"),
        plot.title = element_text(color = "black", face = "bold"),
        
        strip.placement = "outside",
        strip.text.y.left = element_text(angle = 0, face = "bold",
                                         margin = margin(0, 2, 0, 2)),          # CHANGED
        strip.background = element_rect(color = NA),                                      # CHANGED
        strip.switch.pad.grid = unit(0, "pt"),                                  # ADDED
        panel.border = element_blank(),                                         # ADDED
        panel.spacing.y = unit(0.3, "lines"),
        plot.title.position = "plot"
    )

# Build the gtable
g <- ggplotGrob(clus9_heatmap)

# Find left-axis cells (these hold the gene name labels)
axis_idx <- grep("^axis-l", g$layout$name)

# Inject a matching background rect into each axis cell
# axis-l grobs run top-to-bottom, matching strip_fills order
for (i in seq_along(axis_idx)) {
    bg_rect <- rectGrob(gp = gpar(fill = strip_fills[i], col = NA))
    g$grobs[[axis_idx[i]]] <- grobTree(bg_rect, g$grobs[[axis_idx[i]]])
}

# Draw or save
grid.newpage(); grid.draw(g)

# Save as PNG
ggsave(paste0(project_path, "figures/4_monocyte_longitudinal_analysis/clus9_heatmap.png"), g, width = 6, height = 8)
```



