---
title: "5: Generate supplemental spreadsheet"
output: html_notebook
---

In this final notebook, we generate the supplemental Excel workbook accompanying our manuscript. 

We start by setting the project path and loading packages.

```{r}
# Set project path
project_path <- "/Users/juliaelrod/Desktop/cmu/research/scleroderma/jSSc_ASCT_monocytes_manuscript/code_for_manuscript/"

# Load packages
library(dplyr)
library(openxlsx2)
library(Seurat)
library(SeuratDisk)
library(tidyr)
```

We begin by creating the workbook.

```{r}
wb <- wb_workbook()

# Helper: add a bold title/label cell
wb_add_title <- function(wb, sheet, x, start_row, start_col = 1) {
  wb %>%
    wb_add_data(sheet = sheet, x = x, start_row = start_row, start_col = start_col) %>%
    wb_add_font(sheet = sheet,
                dims = wb_dims(rows = start_row, cols = start_col),
                bold = "1")
}

# Helper: add a data frame with bolded headers and (optionally) bolded first column.
# Numeric columns are rounded to 3 significant figures.
wb_add_table <- function(wb, sheet, x, start_row = 1, start_col = 1, bold_first_col = TRUE) {
  # Round numeric columns to 3 significant figures (base R, tolerates blank colnames)
  num_cols <- vapply(x, is.numeric, logical(1))
  x[num_cols] <- lapply(x[num_cols], signif, 3)
  end_col <- start_col + ncol(x) - 1
  wb <- wb %>%
    wb_add_data(sheet = sheet, x = x, start_row = start_row, start_col = start_col) %>%
    wb_add_font(sheet = sheet,
                dims = wb_dims(rows = start_row, cols = start_col:end_col),
                bold = "1")
  if (bold_first_col && nrow(x) > 0) {
    wb <- wb %>%
      wb_add_font(sheet = sheet,
                  dims = wb_dims(rows = (start_row + 1):(start_row + nrow(x)),
                                 cols = start_col),
                  bold = "1")
  }
  wb
}
```

1. Get patient demographic information.

```{r}
# Read in patient meta data
patient_meta_data <- read.csv(paste0(project_path, "data/patient_meta_data.csv"))

# Remove blank rows
patient_meta_data <- patient_meta_data %>% filter(if_all(everything(), ~ . != "" & !is.na(.)))
# Get rid of extra space in Race column
patient_meta_data$Race <- gsub(" ", "", patient_meta_data$Race)
# Rename columns as desired
patient_meta_data <- patient_meta_data %>%
  dplyr::rename(
    "NRCOS ID" = "subject_id",
    "Age at time of ASCT (years)" = "Age.at.time.of.ASCT..years.",
    "Age of onset (years)" = "Age.of.onset..years.",
    "Disease duration at time of ASCT (years)" = "Disease.Duration.at.time.of.ASCT..years.",
    "Disease subtype" = "Disease.Subtype",
    "Autoantibody positivity" = "X....Auto.antibody.positivity",
    "MRSS" = "X....mRSS.",
    "Treatment prior to ASCT" = "Treatment.prior.to.ASCT",
    "Major manifestations" = "Major.manifestations"
  )
```

Get healthy demographics. Add both patient and healthy demographics to sheet.

```{r}
# Read in healthy demographics
healthy_demographics <- read.csv(paste0(project_path, "data/healthy_demographics.csv"))

# Get rid of unnecessary columns and rename variables as desired
healthy_demographics <- healthy_demographics %>% 
  mutate(
  "Sample ID" = c("H1", "H2", "H3", "H4")) %>%
  select(
  "Sample ID",
  "Age" = "age",
  "Sex" = "sex",
  "Race" = "race",
  "Ethnicity" = "ethnicity"
)

# Get sheet #
sheet_num <- length(wb$sheet_names) + 1
# Generate sheet name
sheet_name <- paste0("Table ", sheet_num, "꞉ Demographics")

# Row positions
patient_label_row <- 1
patient_data_row  <- 2
healthy_label_row <- patient_data_row + nrow(patient_meta_data) + 3  # +1 for header, +2 blank rows
healthy_data_row  <- healthy_label_row + 1

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_title(sheet = sheet_name, x = "Patient demographics", start_row = patient_label_row) %>%
  wb_add_table(sheet = sheet_name, x = patient_meta_data,      start_row = patient_data_row) %>%
  wb_add_title(sheet = sheet_name, x = "Healthy demographics", start_row = healthy_label_row) %>%
  wb_add_table(sheet = sheet_name, x = healthy_demographics,   start_row = healthy_data_row)
```

2. Now add top 100 markers and genes used for annotation for each cell type

```{r}
# Load in Seurat object
scler_pbmc <- LoadH5Seurat(paste0(project_path, "data/h5seurat/scler_pbmc_11.h5seurat"))

# Get top 100 RNA markers for each cell type
RNA_markers <- FindAllMarkers(scler_pbmc, assay = "RNA", group.by = "cell_type", only.pos = T) %>%
  group_by(cluster) %>%
  slice_min(order_by = p_val, n = 100, with_ties = FALSE) %>%
  split(~ cluster)

# Get top 20 ADT markers for each cell type
ADT_markers <- FindAllMarkers(scler_pbmc, assay = "DSB", group.by = "cell_type", only.pos = T) %>%
  group_by(cluster) %>%
  slice_min(order_by = p_val, n = 20, with_ties = FALSE) %>%
  split(~ cluster)

# Get cluster #s for each cell type
cluster_nums <- scler_pbmc$cell_type %>% levels() %>% lapply(function(CELLTYPE){scler_pbmc@meta.data %>% filter(cell_type == CELLTYPE) %>% pull(seurat_clusters) %>% unique() %>% sort()}) %>% setNames(levels(scler_pbmc$cell_type))

# Get annotation markers
RNA_annotation <- list(
  "monocyte" = c("FCN1", "S100A8", "S100A9"),
  "B cell" = c("CD79A", "MS4A1", "BANK1"),
  "T/B biphenotypic" = c("CD79A", "MS4A1", "BANK1", "CD3D", "CD3E", "IL7R"),
  "CD4+ T cell" = c("CD3D", "CD3E", "IL7R", "CD8A-", "CD8B-"),
  "CD8+ T cell" = c("CD3D", "CD3E", "IL7R", "CD8A", "CD8B"),
  "natural killer/T cell" = c("STAT4", "RORA"),
  "natural killer" = c("GNLY", "NKG7", "KLRB1"),
  "mDC" = c("CD1C", "CLEC10A"),
  "pDC" = c("IL3RA", "LILRA4", "CLEC4C"),
  "HSPC" = c("CD34", "MYB", "KIT"),
  "platelet" = c("PF4", "PPBP", "TUBB1")
)

ADT_annotation <- list(
  "monocyte" = c("CD14"),
  "B cell" = c("CD19"),
  "T/B biphenotypic" = c("CD19", "CD3"),
  "CD4+ T cell" = c("CD3", "CD4"),
  "CD8+ T cell" = c("CD3", "CD8"),
  "natural killer/T cell" = c("CD3", "CD56", "CD16"),
  "natural killer" = c("CD56", "CD16"),
  "mDC" = c("HLA-DR"),
  "pDC" = c("CD123"),
  "HSPC" = c("CD99"),
  "platelet" = c("CD41")
)

# Get cell types
cell_types <- names(RNA_markers)

# Function to display data as block
make_block <- function(lst, label, n_rows) {
  block <- data.frame(Section = c(label, rep("", n_rows - 1)))
  for (ct in cell_types) {
    vec <- lst[[ct]]
    length(vec) <- n_rows
    block[[ct]] <- vec
  }
  block
}

# Function to display data as single row
make_single_row <- function(lst, label) {
  block <- data.frame(Section = label)
  for (ct in cell_types) {
    block[[ct]] <- paste(lst[[ct]], collapse = ", ")
  }
  block
}

# Extract RNA and ADT genes
RNA_genes <- lapply(RNA_markers, function(d) d$gene)
ADT_genes <- lapply(ADT_markers, function(d) d$gene)   # adjust if ADT uses a different column name

all_cell_df <- rbind(
  make_single_row(cluster_nums,   "Cluster IDs"),
  make_single_row(RNA_annotation, "RNA markers"),
  make_single_row(ADT_annotation, "ADT markers"),
  make_block(RNA_genes, "Top 100 DE genes",      100),
  make_block(ADT_genes, "Top 20 DE antibodies",   20)
)
# Blank out the "Section" header
colnames(all_cell_df)[1] <- ""

# Get sheet #
sheet_num <- length(wb$sheet_names) + 1
# Generate sheet name
sheet_name <- paste0("Table ", sheet_num, "꞉ All cell markers")
# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = all_cell_df)
```

3. Now, we do the same thing for monocyte subtypes. 

```{r}
# Load in Seurat object
monocytes <- LoadH5Seurat(paste0(project_path, "data/h5seurat/monocytes_3.h5seurat"))

# Get top 100 RNA markers for each cell type
RNA_markers <- FindAllMarkers(monocytes, assay = "RNA", group.by = "cell_type", only.pos = T) %>%
  group_by(cluster) %>%
  slice_min(order_by = p_val, n = 100, with_ties = FALSE) %>%
  split(~ cluster)

# Get top 20 ADT markers for each cell type
ADT_markers <- FindAllMarkers(monocytes, assay = "DSB", group.by = "cell_type", only.pos = T) %>%
  group_by(cluster) %>%
  slice_min(order_by = p_val, n = 20, with_ties = FALSE) %>%
  split(~ cluster)

# Get cluster #s for each cell type
monocytes$cell_type <- factor(monocytes$cell_type, unique(monocytes$cell_type))
cluster_nums <- monocytes$cell_type %>% levels() %>% lapply(function(CELLTYPE){monocytes@meta.data %>% filter(cell_type == CELLTYPE) %>% pull(seurat_clusters) %>% unique() %>% sort()}) %>% setNames(levels(monocytes$cell_type))

# Get annotation markers
RNA_annotation <- list(
  "CD14+" = c("CD14"),
  "CD16+" = c("FCGR3A"),
  "B-like" = c("CD79A", "MS4A1"),
  "T-like" = c("CD3E", "CD3D"),
  "NK-like" = c("NKG7", "KLRD1")
)

ADT_annotation <- list(
  "CD14+" = c("CD14"),
  "CD16+" = c("CD16"),
  "B-like" = c("CD19", "IgD"),
  "T-like" = c("CD3"),
  "NK-like" = c("CD56")
)

# Get cell types
cell_types <- names(RNA_markers)


# Extract RNA and ADT genes
RNA_genes <- lapply(RNA_markers, function(d) d$gene)
ADT_genes <- lapply(ADT_markers, function(d) d$gene)   # adjust if ADT uses a different column name

monocyte_df <- rbind(
  make_single_row(cluster_nums,   "Cluster IDs"),
  make_single_row(RNA_annotation, "RNA markers"),
  make_single_row(ADT_annotation, "ADT markers"),
  make_block(RNA_genes, "Top 100 DE genes",      100),
  make_block(ADT_genes, "Top 20 DE antibodies",   20)
)
# Blank out the "Section" header
colnames(monocyte_df)[1] <- ""

# Get sheet #
sheet_num <- length(wb$sheet_names) + 1
# Generate sheet name
sheet_name <- paste0("Table ", sheet_num, "꞉ Monocyte markers")
# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  wb_add_table(sheet = sheet_name, x = monocyte_df)
```

4. Now, let's also add the cell type counts 

```{r}
counts_by_timepoint <- table(scler_pbmc$cell_type,
                       scler_pbmc$group) %>%
  as.data.frame() %>%
  setNames(c("Cell type", "Time point", "Count")) %>%
  pivot_wider(names_from = "Time point", values_from = "Count") %>%
  dplyr::rename("Healthy" = "healthy",
                "jSSc baseline" = "baseline",
                "jSSc 6 months" = "six",
                "jSSc 12 months" = "twelve",
                "jSSc 24 months" = "twen4") %>%
  dplyr::mutate(`Cell type` = as.character(`Cell type`)) %>%
  {rbind(.,
         c("Total by time point", colSums(sapply(.[, -1], as.numeric))),
         c("Total", rep("", ncol(.) - 2), sum(sapply(.[, -1], as.numeric))))}

pct_by_timepoint <- counts_by_timepoint %>%
  dplyr::slice(-dplyr::n()) %>%   # drop the grand "Total" row
  dplyr::mutate(dplyr::across(-`Cell type`, ~ {
    x <- as.numeric(.)
    x / x[length(x)] * 100
  }))

counts_by_library <- table(scler_pbmc$cell_type,
                       scler_pbmc$LIBRARY) %>%
  as.data.frame() %>%
  setNames(c("Cell type", "Library", "Count")) %>%
  pivot_wider(names_from = "Library", values_from = "Count") %>%
    dplyr::rename("Healthy" = "H",
           "NRCOS 515" = "515",
           "NRCOS 591" = "591",
           "NRCOS 594" = "594")%>%
  dplyr::mutate(`Cell type` = as.character(`Cell type`)) %>%
  {rbind(.,
         c("Total by library", colSums(sapply(.[, -1], as.numeric))),
         c("Total", rep("", ncol(.) - 2), sum(sapply(.[, -1], as.numeric))))}

pct_by_library <- counts_by_library %>%
  dplyr::slice(-dplyr::n()) %>%   # drop the grand "Total" row
  dplyr::mutate(dplyr::across(-`Cell type`, ~ {
    x <- as.numeric(.)
    x / x[length(x)] * 100
  }))

# Get sheet #
sheet_num <- length(wb$sheet_names) + 1
# Generate sheet name
sheet_name <- paste0("Table ", sheet_num, "꞉ Cell type counts")

# Column positions
left_col  <- 1
right_col <- ncol(counts_by_timepoint) + 2   # gap of 1 column between left/right tables

# Row positions
upper_label_row <- 1
upper_data_row  <- 2
upper_height    <- max(nrow(counts_by_timepoint), nrow(counts_by_library))
lower_label_row <- upper_data_row + upper_height + 3   # +1 header, +2 blank rows
lower_data_row  <- lower_label_row + 1

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  # Upper left
  wb_add_title(sheet = sheet_name, x = "Raw counts by time point",
               start_row = upper_label_row, start_col = left_col) %>%
  wb_add_table(sheet = sheet_name, x = counts_by_timepoint,
               start_row = upper_data_row,  start_col = left_col) %>%
  # Upper right
  wb_add_title(sheet = sheet_name, x = "Raw counts by library",
               start_row = upper_label_row, start_col = right_col) %>%
  wb_add_table(sheet = sheet_name, x = counts_by_library,
               start_row = upper_data_row,  start_col = right_col) %>%
  # Lower left
  wb_add_title(sheet = sheet_name, x = "Proportions by time point",
               start_row = lower_label_row, start_col = left_col) %>%
  wb_add_table(sheet = sheet_name, x = pct_by_timepoint,
               start_row = lower_data_row,  start_col = left_col) %>%
  # Lower right
  wb_add_title(sheet = sheet_name, x = "Proportions by library",
               start_row = lower_label_row, start_col = right_col) %>%
  wb_add_table(sheet = sheet_name, x = pct_by_library,
               start_row = lower_data_row,  start_col = right_col)
```

5. Same thing for monocyte subtypes.

```{r}
counts_by_timepoint_monocytes <- table(monocytes$cell_type,
                       monocytes$group) %>%
  as.data.frame() %>%
  setNames(c("Cell type", "Time point", "Count")) %>%
  pivot_wider(names_from = "Time point", values_from = "Count") %>%
  dplyr::rename("Healthy" = "healthy",
                "jSSc baseline" = "baseline",
                "jSSc 6 months" = "six",
                "jSSc 12 months" = "twelve",
                "jSSc 24 months" = "twen4") %>%
  dplyr::mutate(`Cell type` = as.character(`Cell type`)) %>%
  {rbind(.,
         c("Total by time point", colSums(sapply(.[, -1], as.numeric))),
         c("Total", rep("", ncol(.) - 2), sum(sapply(.[, -1], as.numeric))))}

pct_by_timepoint_monocytes <- counts_by_timepoint_monocytes %>%
  dplyr::slice(-dplyr::n()) %>%   # drop the grand "Total" row
  dplyr::mutate(dplyr::across(-`Cell type`, ~ {
    x <- as.numeric(.)
    x / x[length(x)] * 100
  }))

counts_by_library_monocytes <- table(monocytes$cell_type,
                       monocytes$LIBRARY) %>%
  as.data.frame() %>%
  setNames(c("Cell type", "Library", "Count")) %>%
  pivot_wider(names_from = "Library", values_from = "Count") %>%
    dplyr::rename("Healthy" = "H",
           "NRCOS 515" = "515",
           "NRCOS 591" = "591",
           "NRCOS 594" = "594")%>%
  dplyr::mutate(`Cell type` = as.character(`Cell type`)) %>%
  {rbind(.,
         c("Total by library", colSums(sapply(.[, -1], as.numeric))),
         c("Total", rep("", ncol(.) - 2), sum(sapply(.[, -1], as.numeric))))}

pct_by_library_monocytes <- counts_by_library_monocytes %>%
  dplyr::slice(-dplyr::n()) %>%   # drop the grand "Total" row
  dplyr::mutate(dplyr::across(-`Cell type`, ~ {
    x <- as.numeric(.)
    x / x[length(x)] * 100
  }))

# Get sheet #
sheet_num <- length(wb$sheet_names) + 1
# Generate sheet name
sheet_name <- paste0("Table ", sheet_num, "꞉ Monocyte type counts")

# Column positions
left_col  <- 1
right_col <- ncol(counts_by_timepoint_monocytes) + 2   # gap of 1 column between left/right tables

# Row positions
upper_label_row <- 1
upper_data_row  <- 2
upper_height    <- max(nrow(counts_by_timepoint_monocytes), nrow(counts_by_library_monocytes))
lower_label_row <- upper_data_row + upper_height + 3   # +1 header, +2 blank rows
lower_data_row  <- lower_label_row + 1

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  # Upper left
  wb_add_title(sheet = sheet_name, x = "Raw counts by time point",
               start_row = upper_label_row, start_col = left_col) %>%
  wb_add_table(sheet = sheet_name, x = counts_by_timepoint_monocytes,
               start_row = upper_data_row,  start_col = left_col) %>%
  # Upper right
  wb_add_title(sheet = sheet_name, x = "Raw counts by library",
               start_row = upper_label_row, start_col = right_col) %>%
  wb_add_table(sheet = sheet_name, x = counts_by_library_monocytes,
               start_row = upper_data_row,  start_col = right_col) %>%
  # Lower left
  wb_add_title(sheet = sheet_name, x = "Proportions by time point",
               start_row = lower_label_row, start_col = left_col) %>%
  wb_add_table(sheet = sheet_name, x = pct_by_timepoint_monocytes,
               start_row = lower_data_row,  start_col = left_col) %>%
  # Lower right
  wb_add_title(sheet = sheet_name, x = "Proportions by library",
               start_row = lower_label_row, start_col = right_col) %>%
  wb_add_table(sheet = sheet_name, x = pct_by_library_monocytes,
               start_row = lower_data_row,  start_col = right_col)
```

6. Longitudinal DEGs in SSc monocytes

```{r}
results_dir <- paste0(project_path, "data/DEG_pathway_CSVs/")
down_res <- read.csv(paste0(results_dir, "down_DEGs.csv"))
# Adjust names + select desired variables
down_res <- down_res %>% select(
  "Gene" = "name",
  "P-value" = "pval",
  "Adjusted p-value" = "adj_pval",
  "Log-fold change" = "lfc"
)
up_res <- read.csv(paste0(results_dir, "up_DEGs.csv"))
# Adjust names + select desired variables
up_res <- up_res %>% select(
  "Gene" = "name",
  "P-value" = "pval",
  "Adjusted p-value" = "adj_pval",
  "Log-fold change" = "lfc"
)
# Get sheet #
sheet_num <- length(wb$sheet_names) + 1
# Generate sheet name
sheet_name <- paste0("Table ", sheet_num, "꞉ long. monocyte DEGs")

# Column positions
left_col  <- 1
right_col <- ncol(down_res) + 2   # gap of 1 column between left/right tables

# Row positions
label_row <- 1
data_row  <- 2

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  # Left
  wb_add_title(sheet = sheet_name, x = "Decreasing DEGs over time since ASCT in jSSc monocytes",
               start_row = label_row, start_col = left_col) %>%
  wb_add_table(sheet = sheet_name, x = down_res,
               start_row = data_row,  start_col = left_col, bold_first_col = FALSE) %>%
  # Right
  wb_add_title(sheet = sheet_name, x = "Increasing DEGs over time since ASCT in jSSc monocytes",
               start_row = label_row, start_col = right_col) %>%
  wb_add_table(sheet = sheet_name, x = up_res,
               start_row = data_row,  start_col = right_col, bold_first_col = FALSE)
```

7. Longitudinal pathways in SSc monocytes 

```{r}
down_pathways <- read.csv(paste0(results_dir, "down_GSEA.csv"))
# Only include pvalue < 0.01 and rename variables
down_pathways <- down_pathways %>%
  filter(pvalue < 0.01) %>%
  select("Pathway ID" = "ID",
         "Fold enrichment" = "FoldEnrichment",
         "P-value" = "pvalue",
         "Adjusted p-value" = "p.adjust",
         "Genes" = "geneID",
         "Number of genes" = "Count")
up_pathways <- read.csv(paste0(results_dir, "up_GSEA.csv"))
# Only include pvalue < 0.01 and rename variables
up_pathways <- up_pathways %>%
  filter(pvalue < 0.01) %>%
  select("Pathway ID" = "ID",
         "Fold enrichment" = "FoldEnrichment",
         "P-value" = "pvalue",
         "Adjusted p-value" = "p.adjust",
         "Genes" = "geneID",
         "Number of genes" = "Count")
# Get sheet #
sheet_num <- length(wb$sheet_names) + 1
# Generate sheet name
sheet_name <- paste0("Table ", sheet_num, "꞉ long. monocyte GSEA")

# Column positions
left_col  <- 1
right_col <- ncol(down_pathways) + 2   # gap of 1 column between left/right tables

# Row positions
label_row <- 1
data_row  <- 2

# Add to workbook
wb <- wb %>%
  wb_add_worksheet(sheet_name) %>%
  # Left
  wb_add_title(sheet = sheet_name, x = "Decreasing pathways over time since ASCT in jSSc monocytes",
               start_row = label_row, start_col = left_col) %>%
  wb_add_table(sheet = sheet_name, x = down_pathways,
               start_row = data_row,  start_col = left_col, bold_first_col = FALSE) %>%
  # Right
  wb_add_title(sheet = sheet_name, x = "Increasing pathways over time since ASCT in jSSc monocytes",
               start_row = label_row, start_col = right_col) %>%
  wb_add_table(sheet = sheet_name, x = up_pathways,
               start_row = data_row,  start_col = right_col, bold_first_col = FALSE)
```


Save workbook.

```{r}
# First get rid of #N/A values
for (sn in wb$get_sheet_names()) {
  df <- wb_to_df(wb, sheet = sn, na.strings = "#N/A")
  wb <- wb_add_data(wb, sheet = sn, x = df, na.strings = "")
}

# Auto-fit column widths on every sheet
for (sn in wb$get_sheet_names()) {
  df <- wb_to_df(wb, sheet = sn)
  if (ncol(df) > 0) {
    wb <- wb_set_col_widths(wb, sheet = sn, cols = 1:ncol(df), widths = "auto")
  }
}

wb_save(wb, paste0(project_path, "data/supplement/JKE_supplement.xlsx"), overwrite = T)
```

